---
title: "Are There Enough Buses?"
editor: visual
---

# which lines are most important to which schools? Which lines are ending up at schools in the morning?

```{r}
school_hours_stops <- final_schedule_object %>% 
  filter(
    hour.y==6|
      hour.y==7|
      hour.y==8|
      (hour.y==14&minute.y>20)|
      (hour.y==15)|(hour.y==16)
           ) 
```

```{r}
# create left-join-able version of the stops object 
D_school_parcels <- st_read("data/shapes/school_parcels/school_parcels.geojson") %>% 
  distinct(.keep_all = T) %>%
  mutate(schools_at_parcel = 
           case_when(
             schools_at_parcel == "Crossroads School, The" ~ "The Crossroads School",
             schools_at_parcel == "Belair-Edison School, The" ~ "The Belair-Edison School",
             schools_at_parcel == "Dallas F. Nicholas, Sr., Elementary School" ~ "Dallas F. Nicholas Sr. Elementary School",
             schools_at_parcel == "Green School Of Baltimore, The" ~ "The Green School Of Baltimore",
             schools_at_parcel == "Reach! Partnership School, The" ~ "The Reach! Partnership School",
             schools_at_parcel == "Dr. Bernard Harris, Sr., Elementary School" ~ "Dr. Bernard Harris Sr. Elementary School",
             schools_at_parcel == "Historic Samuel Coleridge-Taylor Elementary School, The, Joseph C. Briscoe Academy" ~ "The Historic Samuel Coleridge-Taylor Elementary School, The Joseph C. Briscoe Academy",
             TRUE ~ schools_at_parcel
           )) %>% 
  st_transform(4326) %>% 
  st_set_crs(4326)


transformed_df <- data.frame()

# Iterate through each row of the sf object
for (i in 1:nrow(D_school_parcels)) {
  # Split the schools_at_parcel column by comma
  schools <- strsplit(as.character(D_school_parcels[i, "schools_at_parcel"]), ", ")[[1]]
  
  # Create a new row for each school listed
  for (school in schools) {
    # Duplicate the row and update the school_name column
    new_row <- D_school_parcels[i, ]
    new_row$school_name <- school
    
    # Add the new row to the transformed dataframe
    transformed_df <- rbind(transformed_df, new_row)
  }
}

D_school_parcels <- transformed_df 


#create buffer around schools
# this is used to calculate the distance between buses and schools
# "r_2_m" is the distance we use to create the buffer around each school
# "D_school_locs_buffer" is the school data with a buffer around each school
r_2_m <- set_units(.5, "miles") 
D_school_parcels_buff <- st_transform(D_school_parcels, 29902)
D_school_parcels_buffer <-st_buffer(D_school_parcels_buff, r_2_m)


# this gives us a map of all schools with a half mile buffer zone surrounding each 
# the only thing we're doing below is correcting the CRS
D_school_parcels_buffer_all <- st_transform(D_school_parcels_buffer, 4326)

D_school_locs <- st_read("data/shapes/city_schools/Baltimore_City_Schools.shp") %>% 
  st_transform(4326) %>% 
  st_set_crs(4326)




# filter parcels down to high schools: 
hs_locs_join <- 
  D_school_locs %>% 
  filter(str_detect(class, "9 - 12|6 - 12|6 - 8|5 - 8")) %>% 
  mutate(join_addr = str_to_lower(address), 
  ) 


hs_locs_join_df <- 
  hs_locs_join %>% 
  as.data.frame()

hs_parcels_buffer <- 
  D_school_parcels %>%
  st_join(hs_locs_join) %>%
  dplyr::select(school_name, name, join_addr, fulladdr) %>% 
  filter(!is.na(name)) %>% 
  distinct(name, .keep_all = T) %>% 
  as.data.frame() %>% 
  left_join(hs_locs_join_df,
            by = c(name = "name")) %>% 
  filter(str_detect(class, "9 - 12|6 - 12|6 - 8|5 - 8")) %>% 
  st_as_sf() %>%
  st_transform(29902) %>% 
  st_buffer(r_2_m) %>% 
  st_transform(4326) 


# join df for hs parcels (we just need school number from here )
hs_parcel_join <- 
  hs_parcels_buffer %>% 
  as.data.frame() %>% 
  mutate(school_number = as.character(prg_num))

D_stops <- read_csv("data/setup/Baltimore_Region_Public_Transit_(stops).csv") %>% 
  filter(is.na(X)==FALSE, is.na(Y)==FALSE, Jurisdiction == "Baltimore City") %>% 
  st_as_sf(coords = c("X","Y")) %>% 
  st_set_crs(2248) %>%
  st_transform(4326) %>% 
  st_set_crs(4326)


D_school_stop_cens_j<- D_stops %>% 
  dplyr::filter(str_detect(str_to_upper(TYPE_Alt), "BUS")) %>%
  # next join to the map of all schools and their buffer zones
  st_join(hs_parcels_buffer)
  # filter for high schools
  # filter(class%in%c("9 - 12"))%>% 
  # select relevant variables

# for join
D_school_stop_cens_j_join<- D_stops %>% 
  dplyr::filter(str_detect(str_to_upper(TYPE_Alt), "BUS")) %>%
  # next join to the map of all schools and their buffer zones
  st_join(hs_parcels_buffer) %>% 
  mutate(Stop_Number = as.character(Stop_Number)) %>%
  # filter for high schools
  # filter(class%in%c("9 - 12"))%>%
  # select relevant variables 
  dplyr::select(Stop_Number, Stop_Name, school_name, name) %>% 
  as.data.frame()

which_stops_which_school <- 
  D_school_stop_cens_j %>% 
  group_by(Stop_Number) %>% 
  summarise(name = list((name))
            ) %>% 
  as.data.frame()

hs_parcels <- 
    D_school_parcels %>% 
  st_join(hs_locs_join) %>%
  dplyr::select(school_name, name, join_addr, fulladdr) %>% 
  filter(!is.na(name)) %>% 
  distinct(name, .keep_all = T) %>% 
  as.data.frame() %>% 
  left_join(hs_locs_join_df,
            by = c(name = "name")) %>% 
  filter(str_detect(class, "9 - 12|6 - 12|6 - 8|5 - 8")) %>% 
  st_as_sf() %>%
  st_transform(4326) 


```

```{r}

# join with bus stops 
D_bus_locs_hours_near_stops <- 
  school_hours_stops %>% 
  mutate(stop_id = as.character(stop_id)) %>%
  # join to the map of all bus stops and their buffer zones
  left_join(D_school_stop_cens_j_join, by = c("stop_id" = "Stop_Number")) %>% 
  # filter out any bus locations that aren't near a bus stop 
  filter(is.na(name)==FALSE) %>% 
  filter(is.na(school_name)==FALSE)


D_bus_locs_hours_near_stops_school<-
  D_bus_locs_hours_near_stops %>% 
  # turn it into a df 
  data.frame %>% 
  # join to dataframe of bus stops near schools
  # left_join(D_school_stop_cens_j, by = "Stop_Number") %>%
  # mutate(time_reported_et = with_tz(time_reported_et, "America/New_York")) %>% 
  # mutate exact date, and time buckets (either morning or afternoon)
  mutate(
    # date = str_c(year, "-", month, "-", day_of_month) %>% ymd,
         time = case_when(hour.y%in%c(14,15,16)~"afternoon",
                          hour.y%in%c(6,7,8)~"morning"))%>% 
  # join to school enrollment data
  # left_join(school_enrollments, by = c("school_name"="school_name_tracts")) %>% 
  # join to school start and stop times
  left_join(sst_new, by = c("name"="name")) %>% 
  filter(!is.na(am_bell) | !is.na(pm_bell)) %>% 
  ungroup %>% 
  filter(!is.na(hour) | !is.na(minute)) %>% 
  # mutate the exact time of the bus location and turn the bell times into hms objects
  mutate(
    bus_hour = as_hms(timestamp),
    closing_bell = as_hms(pm_bell),
    opening_bell = as_hms(am_bell),
    # create a variable that tells us if the bus location is near a school right before or right after the bell (1 hour)
    keep_filt = 
      case_when(
        time =="afternoon"&
          (bus_hour>closing_bell)&
          (bus_hour<(closing_bell+ as_hms("01:00:00")))~TRUE,
        time=="morning"&
          (bus_hour<opening_bell)&
          (bus_hour>(opening_bell- as_hms("01:00:00")))~TRUE)
    ) %>% 
  # filter out missing times (which presumably happen as a result of broken sensors?)
  filter(is.na(timestamp)==FALSE) %>% 
  # filter out times we don't care about 
  filter(is.na(keep_filt)==FALSE) %>% 
  st_as_sf() %>%
  # join to hs parcel
  st_join(hs_parcels_buffer) %>%
  # make sure the name of the school associated with the bus stop is the same as the name of the school associated with the buffer zone
  filter(name.x == name.y) %>%
  dplyr::select(-name.y) %>% 
  rename(name = name.x) 






```

From there we count total number of buses that passed through each school's buffer zone on a given day

```{r}
  # start with the df we just created 
D_bus_locs_hours_near_stops_school_2<-D_bus_locs_hours_near_stops_school%>% 
  # filter for buses that arrived right before or right after school
  filter(is.na(keep_filt)==FALSE) %>% 
  # group by year, month, day, hour, school, and bus number
  group_by(service_date, hour.y, name, id, 
           # route, 
           trip_id, school_number) %>% 
  # count the number of buses that arrived right before or right after school
  summarise(n = n()) %>% 
  ungroup %>%
  # group by year, month, day, hour, and school
  group_by(service_date, hour.y, name, school_number) %>% 
  summarise(n = n()) 
# final DF is a list of the number of buses that arrived right before or right after school at each school on each day

```

Here, we're adding some additional context to the data. We're adding the exact date and time buckets (either morning or afternoon) to the data.

We also re-join start/stop times

```{r}
# begin with the df we just created
D_bus_locs_hours_near_stops_school_3<-D_bus_locs_hours_near_stops_school_2 %>%
  # join to start/stop times for additional context
  left_join(sst_new, by = c("name"="name")) %>% 
  # mutate the exact date and time buckets (either morning or afternoon)
  # mutate(hour = as_factor(hour))
  mutate(
    # date = str_c(year, "-",month,"-",day_of_month) %>% ymd,
         time = case_when(hour.y%in%c(14,15,16)~"afternoon",
                          hour.y%in%c(6,7,8)~"morning")
  )

D_bus_locs_hours_near_stops_school_3 %>% 
  # filter(name == "Mergenthaler Vocational-Technical High") %>%
  mapview(zcol = "name", legend = F) + 
  mapview(hs_parcels_buffer, col.region = "yellow", zcol = "name", legend = F)
  
  # join to school enrollment data
  # left_join(school_enrollments, by = c("name"="school_name_tracts"))
# final df is much the same as "D_bus_locs_hours_near_stops_school_2" but with additional context


```

Now we're going to group by school, date, and time of day. We'll calculate the average number of buses that arrived right before or right after school at each school on each day.

```{r}
#holidays and weekends to remove
all_holidays <- 
  c("2017-09-04", "2017-09-21", "2017-09-22", "2017-10-09",
    "2017-11-10", "2017-11-23", "2017-11-24", "2017-12-22",
    "2017-12-25", "2017-12-26", "2017-12-27", "2017-12-28",
    "2017-12-29", "2018-09-03", "2018-09-10", "2018-09-19",
    "2018-09-20", "2018-11-12", "2018-11-22", "2018-11-23",
    "2018-12-24", "2018-12-25", "2018-12-26", "2018-12-27",
    "2018-12-28", "2018-12-31", "2019-09-02", "2019-09-30",
    "2019-10-01", "2019-10-09", "2019-10-14", "2019-11-11",
    "2019-11-27", "2019-11-28", "2019-11-29", "2019-12-23",
    "2019-12-24", "2019-12-25", "2020-01-01", "2020-01-20",
    "2020-02-17", "2020-04-10", "2020-05-25", "2020-06-05",
    "2020-06-11", "2020-06-12", "2020-06-15", "2020-06-16",
    "2020-06-17", "2020-06-18", "2020-06-19", "2021-01-01",
    "2021-01-18", "2021-02-15", "2021-04-02", "2021-05-31",
    "2021-06-18", "2021-06-21", "2021-06-22", "2021-06-23",
    "2021-06-24", "2021-06-25", "2022-01-17", "2022-02-21",
    "2022-04-15", "2022-05-30", "2022-06-17", "2022-06-20",
    "2022-06-21", "2022-06-22", "2022-06-23", "2022-06-24",
    "2022-09-05", "2022-11-11", "2022-11-24", "2022-11-25",
    "2023-01-01", "2023-01-07", "2023-01-08", "2023-01-14",
    "2023-01-15", "2023-01-21", "2023-01-22", "2023-01-28",
    "2023-01-29", "2023-02-04", "2023-02-05", "2023-02-11",
    "2023-02-12", "2023-02-18", "2023-02-19", "2023-02-25",
    "2023-02-26", "2023-03-04", "2023-03-05", "2023-03-11",
    "2023-03-12", "2023-03-18", "2023-03-19", "2023-03-25",
    "2023-03-26", "2023-04-01", "2023-04-02", "2023-04-08",
    "2023-04-09", "2023-04-15", "2023-04-16", "2023-04-22",
    "2023-04-23", "2023-04-29", "2023-04-30", "2023-05-06",
    "2023-05-07", "2023-05-13", "2023-05-14", "2023-05-20",
    "2023-05-21", "2023-05-27", "2023-05-28", "2023-06-03",
    "2023-06-04", "2023-06-10", "2023-06-11", "2024-09-02",
    "2024-10-18", "2024-11-04", "2024-11-05"
    ) %>% 
  ymd() %>% as_date

breaks <- c(ymd("2017-03-24"):ymd("2017-04-03"), ymd("2017-06-16"):ymd("2017-09-05"),
            ymd("2018-03-30"):ymd("2018-04-06"), ymd("2018-06-15"):ymd("2018-09-04"),
            ymd("2019-04-19"):ymd("2019-04-26"), ymd("2019-06-14"):ymd("2019-09-03"),
            ymd("2020-04-06"):ymd("2020-04-13"), ymd("2020-06-15"):ymd("2020-09-08"),
            ymd("2021-03-26"):ymd("2021-04-05"), ymd("2021-06-21"):ymd("2021-09-08"),
            ymd("2022-04-15"):ymd("2022-04-22"), ymd("2022-06-20"):ymd("2022-09-06"),
            ymd("2022-12-22"):ymd("2023-01-04"), ymd("2024-11-27"):ymd("2024-11-30")
            ) %>% 
  as_date()

to_rm <- c(all_holidays, breaks) %>% as_date



# begin with the df we just created
D_bus_locs_timep_near_school<-D_bus_locs_hours_near_stops_school_3 %>%
  # group by school, date, and time (here, we group by time of day rather than hour)
  group_by(name, service_date, time, school_number.x, school_number.y) %>% 
  # count the number of buses that arrived right before or right after school
  summarise(
    n = sum(n)
    # enroll = Enrollment.x[1]
  ) 
  # mutate exact date
  # mutate(date = str_c(year, "-",month,"-",day_of_month) %>% ymd) %>% 
  # filter out days with no school 
  # filter(!(service_date %in% to_rm)) %>%
  # filter out missing times 
  # filter(is.na(time)==FALSE)

```

```{r}
# finally, we group by school and time of day and calculate the average number of buses that arrived right before or right after school
D_school_times_num_bus_avg<-
  D_bus_locs_timep_near_school%>% 
  group_by(name, time, school_number.x, school_number.y) %>% 
  summarise(
    avg_n = mean(n),
    max_n = max(n),
    min_n = min(n),
    median_n = median(n),
    sd_n = sd(n),
    min_day = min(as.Date(service_date)),
    max_day = max(as.Date(service_date)),
    sd_as_pct_of_total_stops = sd(n)/mean(n)*100
    # enroll = enroll[1]
  ) %>% 
  pivot_wider(
    names_from = time,
    values_from = c(avg_n, max_n, min_n, median_n, sd_n, min_day, max_day, sd_as_pct_of_total_stops)
  ) %>% 
  group_by(name) %>% 
  summarise(across(everything(), ~mean(.x, na.rm = T)))  %>% 
  dplyr::select(-school_number.x, -school_number.y, -max_day_afternoon, -min_day_afternoon) %>%
  rename(max_date = max_day_morning,
         min_date = min_day_morning) %>%
  dplyr::select(name,  min_date, max_date, everything()) %>% 
  as.data.frame() 
  # dplyr::select(-where_is_bus_stopped)

D_school_times_num_bus_avg %>% 
  st_drop_geometry(geometry.x) %>% 
  dplyr::select(-geometry.x) %>% 
  write_csv("data/output/bus_arrivals_by_school_through_10_02.csv")





```

```{r}

arrivals_by_line <- D_bus_locs_hours_near_stops_school%>% 
  # filter for buses that arrived right before or right after school
  filter(is.na(keep_filt)==FALSE) %>% 
  mutate(day_of_month = lubridate::day(timestamp),
         month = lubridate::month(timestamp),
         year = lubridate::year(timestamp)) %>%
  # group by year, month, day, hour, school, and bus number
  group_by(year, month, day_of_month,name, id, time,
           route_id,
           trip_id, school_number) %>% 
  # count the number of buses that arrived right before or right after school
  summarise(n = n()) %>% 
  ungroup %>%
  # group by year, month, day, hour, and school
  group_by(
    service_date,
    # year, month, day_of_month, 
           name, time,
           route_id,
           school_number) %>% 
  # summarise(n = n()) %>% 
    summarise(
    avg_n = mean(n),
    max_n = max(n),
    min_n = min(n),
    median_n = median(n),
    sd_n = sd(n),
    min_day = min(as.Date(service_date)),
    max_day = max(as.Date(service_date)),
    sd_as_pct_of_total_stops = sd(n)/mean(n)*100
    # enroll = enroll[1]
  ) %>% 
  left_join(mta_routes, by = c("route_id" = "id")) %>% 
  dplyr::select(route_id, short_name, long_name, everything(), -color, -text_color, -type) 


arrivals_by_line

arrivals_by_line %>% 
  filter(str_detect(name.x, "Western")) %>% 
  mapview(zcol = "short_name")
```

On which lines are students starting their trips?

```{r}

student_tract_summaries <- 
  student_home_tracts_23 %>% 
  filter(str_detect(grades, "9 - 12|6 - 12|6 - 8|5 - 8")) %>% 
  group_by(census_tract) %>% 
  summarise(students = sum(students),
            n_schools = n_distinct(name),
            school_names = list(name),
            tract_point = unique(tract_point))


routes_tracts_crosswalk <- school_list %>% 
  filter(!is.na(route)) %>% 
  group_by(from_id, to_id) %>% 
  filter(segment == min(segment)) %>% 
  mutate(route = as.character(route)) %>%
  left_join(all_routes, by = c("route" = "id")) %>% 
  ungroup() %>% 
  group_by(route, short_name, from_id) %>% 
  summarise(n = n()) %>% 
  mutate(from_id = str_remove_all(from_id, "x")) %>% 
  left_join(student_tract_summaries, by = c("from_id" = "census_tract")) %>%
  dplyr::select(route, short_name, from_id, n, students, n_schools, school_names, tract_point) %>% 
  arrange(from_id) %>% 
  group_by(from_id) %>% 
  mutate(pct_routes = n / sum(n) * 100) %>%
  rename(n_routes = n) %>% 
  ungroup() %>%
  dplyr::select(from_id, route, short_name, n_routes, pct_routes, students, n_schools, school_names, tract_point) 

routes_tracts_crosswalk %>% 
  st_as_sf(crs = 4326) %>%
  mapview()
```
