---
title: "Interactive routes"
editor: visual
---

## Quarto

```{r}
library(devtools)
library(elevatr)
library(r5r)
library(sf)
library(osmextract)
library(rJava)
library(tidyverse)
library(geobr)
library(here)
library(tigris)
library(tidycensus)
library(mapview)
library(RColorBrewer)
library(readxl)
library(janitor)
library(httr)
library(hms)
library(raster)
library(terra)
library(osmdata)

# turn off scientific notation
options(scipen = 999)
```

# Load school shapes

```{r}
# school points 
D_school_locs <- st_read("data/shapes/city_schools/Baltimore_City_Schools.shp") %>% 
  st_transform(4326) %>% 
  st_set_crs(4326)


# filter points down to high schools and middle schools 
hs_locs_join <- 
  D_school_locs %>% 
  filter(str_detect(class, "9 - 12|6 - 12|6 - 8|5 - 8")) %>% 
  mutate(join_addr = str_to_lower(address), 
  ) 

# df version for joining
hs_locs_join_df <- 
  hs_locs_join %>% 
  as.data.frame() %>% 
  mutate(prg_num = as.character(prg_num))

# read in school parcels 
D_school_parcels <- st_read("data/shapes/school_parcels/school_parcels.geojson") %>% 
  distinct(.keep_all = T) %>%
  mutate(schools_at_parcel = 
           case_when(
             schools_at_parcel == "Crossroads School, The" ~ "The Crossroads School",
             schools_at_parcel == "Belair-Edison School, The" ~ "The Belair-Edison School",
             schools_at_parcel == "Dallas F. Nicholas, Sr., Elementary School" ~ "Dallas F. Nicholas Sr. Elementary School",
             schools_at_parcel == "Green School Of Baltimore, The" ~ "The Green School Of Baltimore",
             schools_at_parcel == "Reach! Partnership School, The" ~ "The Reach! Partnership School",
             schools_at_parcel == "Dr. Bernard Harris, Sr., Elementary School" ~ "Dr. Bernard Harris Sr. Elementary School",
             schools_at_parcel == "Historic Samuel Coleridge-Taylor Elementary School, The, Joseph C. Briscoe Academy" ~ "The Historic Samuel Coleridge-Taylor Elementary School, The Joseph C. Briscoe Academy",
             TRUE ~ schools_at_parcel
           )) %>% 
  st_transform(4326) %>% 
  st_set_crs(4326)

# D_school_parcels %>% 
#   st_write("/Users/greg.morton/Documents/baltimoreBanner/Projects/2024/high_school_transportation_routing/data/shapes/school_parcels/parcels_01_22.geojson")


# join school parcels to high school locations
hs_parcels_buffer <- 
  D_school_parcels %>% 
  st_join(hs_locs_join) %>%
  dplyr::select(school_name, name, join_addr, fulladdr) %>% 
  filter(!is.na(name)) %>% 
  distinct(name, .keep_all = T) %>% 
  as.data.frame() %>% 
  left_join(hs_locs_join_df,
            by = c(name = "name")) %>% 
  filter(str_detect(class, "9 - 12|6 - 12|6 - 8|5 - 8")) %>% 
  st_as_sf() %>%
  st_transform(29902)


# join df for hs parcels (we just need school number from here)
hs_parcel_join <- 
  hs_parcels_buffer %>% 
  as.data.frame() %>% 
  mutate(school_number = as.character(prg_num)) %>% 
  ungroup()

# get start stop times 
start_stop_times <- read_csv("data/setup/start_stop_times.csv") %>% 
  janitor::clean_names() %>% 
  mutate(school_number = as.character(school_number))

# select relevant columns 
sst_new <- start_stop_times %>% 
  left_join(hs_parcel_join, by = c("school_number" = "school_number")) %>% 
  filter(!is.na(name)) %>% 
  dplyr::select(school_number, name, am_bell, pm_bell, district_early_release_time, address.y
                ) 

# load functions I need to access MTA API
source("functions/mta_api_functions.R")


```

### Load Baltimore City Shape

```{r}
# get baltimmore city census tracts 
bmore_tracts <- tigris::tracts(
  state = "MD", 
  county = "Baltimore City"
) %>%
  tigris::erase_water() %>%
  st_transform(4326) %>%
  # First make valid
  st_make_valid() %>%
  # Remove any empty geometries
  filter(!st_is_empty(.)) %>%
  # Extract polygons
  st_collection_extract("POLYGON") %>%
  # Remove any potential duplicate vertices and fix other issues
  st_buffer(0) %>%
  # Group and union
  group_by(GEOID, NAME) %>% 
  summarise(geometry = st_union(geometry), .groups = "drop")

# add parks so we can remove them 
parks <- st_read("data/shapes/Parks/Parks.shp") %>% st_transform(32618) 

# add a slight buffer to the parks so we can remove them from the tracts
parks <- st_buffer(parks, 50) %>% st_union() %>% st_transform(4326)

# add schools so we can remove them 
schools <- hs_parcels_buffer %>% 
  st_buffer(100) %>% 
  st_union() %>% 
  st_transform(4326)


grid <- raster("data/raw/GHS_BUILT_S_E2030_GLOBE_R2023A_54009_100_V1_0_R5_C12/GHS_BUILT_S_E2030_GLOBE_R2023A_54009_100_V1_0_R5_C12.tif")

grid <- projectRaster(grid, crs = st_crs(4326)$proj4string)

bbox <- extent(st_bbox(bmore_tracts))

area_bbox <- st_bbox(bmore_tracts) %>% 
  st_as_sfc() %>% 
  st_transform(4326) %>% 
  st_buffer(3000) %>% 
  st_bbox() %>% 
  st_as_sfc() %>% 
  st_transform(4326) %>% 
  st_bbox()

# area_bbox %>% 
#   mapview() +
#   map
```

```{r}
# pbf_file <- "/Users/greg.morton/Documents/baltimoreBanner/Projects/2024/high_school_transportation_routing/data/poa/2025-01-13/baltimore.osm.pbf"

# Read the PBF file using sf
# streets <- sf::st_read(pbf_file, layer = "lines", quiet = TRUE)
# 
# # Filter for roads (you can modify these conditions based on your needs)
# roads <- streets[streets$highway %in% c("motorway", "trunk", "primary", "secondary",
#                                       "tertiary", "residential", "unclassified"), ]
# #   
# # Create an interactive visualization
# map <- mapview(roads,
#               zcol = "highway",  # Color by road type
#               col.regions = rainbow(length(unique(roads$highway))),
#               layer.name = "Street Network",
#               map.types = c("OpenStreetMap", "Esri.WorldImagery"),
#               legend = TRUE,
#               popup = TRUE)
# 

# industrial zones along the water 
mizod <- st_read("/Users/greg.morton/Documents/baltimoreBanner/Projects/2024/high_school_transportation_routing/data/shapes/MIZOD/MIZOD.shp") %>% 
  st_union() %>% 
  st_transform(4326) 




```

```{r}
cropped_raster <- crop(grid, bbox)

baltimore_grid <- rasterToPoints(cropped_raster, spatial = TRUE)

# get Baltimore elevation data from elevatr
# Convert bbox to sf object for get_elev_raster
aoi <- st_as_sfc(st_bbox(area_bbox)) %>%
  st_sf() %>%
  st_set_crs(4326)

# Step 2: Get elevation data using elevatr
# zoom = 9-14 determines resolution (higher = more detailed but larger file)
elevation_raster <- get_elev_raster(locations = aoi, z = 12, clip = "locations")

elevation_terra <- rast(elevation_raster)  # Convert to terra format

writeRaster(elevation_terra, 
            filename = "data/poa/elevation.tif",
            overwrite = TRUE)

```

```{r}
# version of tracts with no parks 
bmore_tracts_no_parks <- 
  bmore_tracts %>% 
  st_difference(parks) %>%
  st_difference(schools) %>%
  st_difference(mizod) 


# get a new grid sample
# baltimore_grid_sampled <-
#   baltimore_grid %>%
#   as.data.frame() %>%
#   st_as_sf(coords = c("x", "y"), crs = 4326) %>%
#   st_difference(parks) %>%
#   st_difference(schools) %>%
#   st_difference(mizod) %>%
#   st_join(bmore_tracts) %>%
#   filter(!is.na(GEOID)) %>%
#   # filter(!is.na(median_income)) %>%
#   group_by(GEOID) %>%
#   filter(GHS_BUILT_S_E2030_GLOBE_R2023A_54009_100_V1_0_R5_C12 > 3000) %>%
#   # Calculate points_to_sample based on population proportion
#   # Use group_modify to sample points
#   # group_modify(~ slice_sample(.x, n = first(.x$points_to_sample), replace = TRUE)) %>%
#   # sample one point from each tract
#   group_modify(~ slice_sample(.x, n = 1, replace = TRUE)) %>%
#   ungroup() %>%
#   st_as_sf(crs = 4326) %>%
#   st_make_valid() %>%
#   mutate(geometry = st_make_valid(geometry)) %>%
#   mutate(lon = map_dbl(geometry, ~st_coordinates(.)[1]),
#          lat = map_dbl(geometry, ~st_coordinates(.)[2])) %>%
#   ungroup() %>%
#   mutate(id = row_number())

baltimore_grid_sampled <-
  st_read("data/shapes/bmore_grid_sampled_01_27.geojson") %>%
  st_as_sf() %>% 
  mutate(key = 1)


# broken_start <-
#   sht_bell_test %>%
#   .[4140,] %>%
#   dplyr::select(geometry) %>%
#   st_as_sf()
# 
baltimore_grid_sampled %>%
  mapview(zcol = "GHS_BUILT_S_E2030_GLOBE_R2023A_54009_100_V1_0_R5_C12", legend = F) +
mapview(bmore_tracts_no_parks) 

```

```{r}
# make a join version with lat and lon for r5r
bmore_tracts_join <-
  bmore_tracts %>% 
  clean_names() %>% 
  dplyr::select(geoid, name, geometry) %>% 
  rename(tract_name = name) %>%
  st_as_sf() %>% 
  st_transform(4326) %>% 
  st_difference(parks) %>% 
  st_make_valid() %>%
  mutate(
    valid_geometry = st_make_valid(geometry),  # Ensure geometries are valid
    tract_point = st_centroid(valid_geometry),  # Calculate the centroid of the land area,
    tract_point_lat = st_coordinates(tract_point)[, 2],  # Extract the latitude of the centroid
    tract_point_lon = st_coordinates(tract_point)[, 1]  # Extract the longitude of the centroid
  ) 


```

### Load student location data

The 'student_home_tract_XX' objects hold info about where a student is going and where they are coming from.

There is a row for each combination of home census tract and school in the data.

```{r}
# read student home tract data 
student_home_tracts_23 <- 
  read_xlsx("data/home_tracts/REVISED Banner Request-Enroll by Prog and Tract-2024.08.08.xlsx", sheet = 2, skip = 5) %>% 
  clean_names() %>% 
  mutate(prog_no = as.character(prog_no)) %>%
  # change all colnames to character
  mutate(across(-c(prog_no, program, enrollment), ~ as.numeric(as.character(.)))) %>%
  # Pivot longer to convert census tract columns into rows
  pivot_longer(
    cols = -c(prog_no, program, enrollment),
    names_to = "census_tract",
    values_to = "students"
  ) %>% 
  left_join(hs_parcel_join, by = c("prog_no" = "school_number")) %>%
  dplyr::select(prog_no, program, enrollment, school_name, name, everything()) %>%
  mutate(students = ifelse(is.na(students), 5, students)) %>%
  filter(!is.na(school_name),
         students > 0
         )  %>%
  dplyr::select(-x, -y, -join_addr.y, -GlobalID, -class, -category, -OBJECTID, -join_addr.x) %>% 
  rename(school_polygon = geometry.x,
         school_point = geometry.y) %>% 
  mutate(school_point = st_make_valid(school_point)) %>% 
  mutate(school_point_lat = st_coordinates(school_point)[, 2],
         school_point_lon = st_coordinates(school_point)[, 1],
         census_tract = str_remove_all(census_tract, "x")) %>% 
  as.data.frame() %>%
  # There are two ways that we can assign students home tracts
  # first we can simply use "bmore_tracts_join", an sf object that simply takes the centerpoint of each tract
  # the upside here is that it's easier and its consistent
  # The downside is that sometimes it can put our point at a place we know no one lives, like a park or industrial area
  # left_join(bmore_tracts_join, by = c("census_tract" = "geoid")) 
  # The second method is to sample human settlement points from the raster grid and join them to the tracts
  # We'll then take a random point from each tract to represent the students home location
  # This is probably more accurate but it's also more computationally expensive
  # We'll do both and compare the results
  left_join(baltimore_grid_sampled, by = c("census_tract" = "GEOID")) %>% 
  rename(
    tract_point_lat = lat,
    tract_point_lon = lon
  ) 




# make enrollment df
enrollment_23 <- 
  read_xlsx("data/home_tracts/REVISED Banner Request-Enroll by Prog and Tract-2024.08.08.xlsx", sheet = 2, skip = 5) %>% 
  clean_names() %>% 
  mutate(prog_no = as.character(prog_no)) %>%
  # change all colnames to character
  mutate(across(-c(prog_no, program, enrollment), ~ as.numeric(as.character(.)))) %>%
  # Pivot longer to convert census tract columns into rows
  pivot_longer(
    cols = -c(prog_no, program, enrollment),
    names_to = "census_tract",
    values_to = "students"
  ) %>% 
  # left_join(hs_parcel_join, by = c("prog_no" = "school_number")) %>%
  dplyr::select(prog_no, program, enrollment, everything()) %>% 
  mutate(students = ifelse(is.na(students), 5, students)) %>%
  filter(!is.na(program),
         students > 0
         )  %>% 
  dplyr::select(
    prog_no,
    program,
    enrollment
  ) %>% 
  mutate(program_join = str_to_lower(str_replace_all(program, " ", "_"))) %>% 
  group_by(program) %>% 
  distinct() %>% 
  ungroup() %>% 
  left_join(hs_locs_join_df, by = c("prog_no" = "prg_num")) %>% 
  filter(!is.na(name))


# get enrollment by tract 
enrollment_by_tract <-
  student_home_tracts_23 %>%
  dplyr::select(
    census_tract,
    name,
    students,
    zipcode
  )




```

# Every possible tract/school combination

```{r}
# create a version of `hs_parcel_join` that gives us a many to many join
hs_parcel_join_m_2_m <- 
  hs_parcel_join %>% 
  mutate(key = 1)

# population by tract 
pop_by_tract <- 
  get_acs(
    year = 2023,
    geography = "tract", 
    variables = "B01003_001", 
    state = "MD", 
    county = "Baltimore City") 


baltimore_grid_sampled_pop <- 
  baltimore_grid_sampled %>% 
  mutate(key = 1) %>% 
  left_join(pop_by_tract, by = c("GEOID" = "GEOID")) %>% 
  rename(pop = estimate) %>% 
  filter(pop > 0)


# create a many to many join
every_missing_tract_school_combo <- 
  baltimore_grid_sampled_pop %>% 
  left_join(hs_parcel_join_m_2_m, by = "key") %>% 
  rename(census_tract = GEOID) %>% 
  anti_join(student_home_tracts_23, by = c("census_tract" = "census_tract", "name" = "name")) %>% 
    as.data.frame() %>% 
    rename(school_polygon = geometry.x,
           school_point = geometry.y) %>% 
  mutate(school_point = st_make_valid(school_point)) %>% 
  mutate(school_point_lat = st_coordinates(school_point)[, 2],
         school_point_lon = st_coordinates(school_point)[, 1],
         census_tract = str_remove_all(census_tract, "x")) %>% 
  as.data.frame() %>% 
  rename(
  tract_point_lat = lat,
  tract_point_lon = lon
)


every_tract_school_combo %>% 
  group_by(GEOID) %>% 
  count()

every_missing_tract_school_combo %>% 
  dplyr::select(
    census_tract,
    tract_point,
    tract_point_lat,
    tract_point_lon,
    school_name,
    school_polygon,
    school_point,
    school_point_lat,
    school_point_lon
  ) %>% 
  View()


# get population by tract so we can filter empty ones 



```

### Load MTA stuff

```{r}
# mta routes 
routes_url <- "https://api.goswift.ly/info/mta-maryland/routes"

# light rail url for api call
light_rail_url <- "https://api.goswift.ly/info/mta-maryland-light-rail/routes/"

# metro url for api call
metro_url <- "https://api.goswift.ly/info/mta-maryland-metro/routes/"

# get bus api response
routes_response <- VERB("GET", routes_url, add_headers('Authorization' = '0352517c0285694758568f98b6f013ba'), content_type("application/octet-stream"), accept("application/json"))

# light rail api response
light_rail_response <- VERB("GET", light_rail_url, add_headers('Authorization' = '0352517c0285694758568f98b6f013ba'), content_type("application/octet-stream"), accept("application/json"))

# metro api response 
metro_response <- VERB("GET", metro_url, add_headers('Authorization' = '0352517c0285694758568f98b6f013ba'), content_type("application/octet-stream"), accept("application/json"))

# routes json
routes_json <- content(routes_response, "text")

# lr json
light_rail_json <- content(light_rail_response, "text")

# metro json
metro_json <- content(metro_response, "text")

# turn em all into dataframes
mta_routes <- 
  jsonlite::fromJSON(routes_json, simplifyDataFrame = T, flatten = F) %>% 
  .$data %>% 
  .$routes %>% 
  clean_names

light_rail_routes <-
    jsonlite::fromJSON(light_rail_json, simplifyDataFrame = T, flatten = F) %>% 
  .$data %>% 
  .$routes %>% 
  clean_names

metro_routes <-
    jsonlite::fromJSON(metro_json, simplifyDataFrame = T, flatten = F) %>%
  .$data %>%
  .$routes %>%
  clean_names

all_routes <- 
  bind_rows(mta_routes, light_rail_routes, metro_routes) %>% 
  mutate(short_name = str_remove_all(short_name, " "))

D_school_parcels <- st_read("data/shapes/school_parcels/school_parcels.geojson") %>% 
  distinct(.keep_all = T) %>%
  mutate(schools_at_parcel = 
           case_when(
             schools_at_parcel == "Crossroads School, The" ~ "The Crossroads School",
             schools_at_parcel == "Belair-Edison School, The" ~ "The Belair-Edison School",
             schools_at_parcel == "Dallas F. Nicholas, Sr., Elementary School" ~ "Dallas F. Nicholas Sr. Elementary School",
             schools_at_parcel == "Green School Of Baltimore, The" ~ "The Green School Of Baltimore",
             schools_at_parcel == "Reach! Partnership School, The" ~ "The Reach! Partnership School",
             schools_at_parcel == "Dr. Bernard Harris, Sr., Elementary School" ~ "Dr. Bernard Harris Sr. Elementary School",
             schools_at_parcel == "Historic Samuel Coleridge-Taylor Elementary School, The, Joseph C. Briscoe Academy" ~ "The Historic Samuel Coleridge-Taylor Elementary School, The Joseph C. Briscoe Academy",
             TRUE ~ schools_at_parcel
           )) %>% 
  st_transform(4326) %>% 
  st_set_crs(4326) %>% 
  mutate(school_number = as.character(school_number)) %>% 
  left_join(enrollment_23, by = c("school_number" = "prog_no")) %>% 
  filter(!is.na(program_join)) 

```

```{r}
### Setup r5r (to fill missing routes we're going to use the GTFS file from 1/27)


# start rjava and allocate enough memory for r5r
rJava::.jinit()
options(java.parameters = "-Xmx50g")
rJava::.jcall("java.lang.System", "S", "getProperty", "java.version")
# un-comment for session details 
# session_info()
# r5r::r5r_sitrep()

# load baltimore shape un-comment if you need to load osm file
# baltimore_shape <-
#   tigris::places("Baltimore", state = "MD", class = "city") %>%
#   filter(NAME == "Baltimore") %>%
#   st_as_sf()  %>%
#   st_transform(4326) 

# define the modes of transit we need gtfs files for
modes <- c("light-rail", "local-bus", "metro")

# get todays date in the format we need
# today <- lubridate::today() %>% 
#   as.character() %>%
#   str_split(" ") %>% 
#   map_chr(1)

# create a directory for the data
# dir.create(here::here("data/poa/", today), showWarnings = F)


# grab today's GTFS files 
# lapply(modes, function(x) {
#   gtfs_url <- paste0("https://feeds.mta.maryland.gov/gtfs/", x)
#   download.file(gtfs_url, paste0("data/poa/",today, "/", x, "_gtfs.zip"))
#   })

# grab .tif elevation 
# tst <- st_read("/Users/greg.morton/Documents/baltimoreBanner/Projects/2024/high_school_transportation/data/NED10M_SpatialMetadata/NED10M_fe2947_October2018.shp") %>% 
#   st_transform(4326) %>% 
#   st_set_crs(4326) 

# get osm data for baltimore
# oe_get("us/maryland", stringsAsFactors = FALSE, quiet = FALSE)
# oe_download_directory()

# get the bounding box of the baltimore shape
# area_bbox <- sf::st_bbox(baltimore_shape)

# path to osmosis
osmosis_path <- "/opt/homebrew/bin/osmosis"

# path to the large .pbf file
large_pbf_path <- "data/poa/maryland-latest.osm.pbf"


# path where you want to save the smaller .pbf file
smaller_pbf <-"data/poa/baltimore_complete_ways_complete_relations.osm.pbf"

# prepare call to osmosis
osmosis_cmd <- sprintf(
  "%s --read-pbf file=%s --bounding-box completeWays=yes left=%s bottom=%s right=%s top=%s --write-pbf file=%s",
  osmosis_path, large_pbf_path,
  area_bbox["xmin"], area_bbox["ymin"], area_bbox["xmax"], area_bbox["ymax"],
  smaller_pbf
)

# call to osmosis
system(osmosis_cmd, intern = TRUE)

# copy the OSM PBF files from data/osm_pbf into today's poa directory
path = paste0("data/poa/", as.character(today), "/baltimore.osm.pbf")

file.copy(
  from = "data/poa/baltimore_complete_ways_complete_relations.osm.pbf",
  to = paste0("data/poa/", as.character(today), "/baltimore.osm.pbf"),
  overwrite = TRUE
)

# copy .tif file 
file.copy(
  from = "data/poa/elevation.tif",
  to = paste0("data/poa/", as.character(today), "/elevation.tif"),
  overwrite = TRUE
)



# setup r5r
r5r_core <- setup_r5(here::here(paste0("data/poa/", "2025-01-27")),
                     temp_dir = F,
                     verbose = T,
                     overwrite = F
                     )

r5r_core

# stop r5r
r5r::stop_r5(r5r_core)
# rJava::.jgc(R.gc = TRUE)
```
```{r}
# get the rest of our school info with a join 
school_info_join <- 
  student_home_tracts_23 %>% 
    dplyr::select(prog_no,
                  program,
                  # name,
                  # bldg_num,
                  # grades,
                  # type
                  # coord_type,
                  # lat,
                  # lon,
    ) %>% 
  distinct()

# Define modes of transit
mode <- c("WALK", "TRANSIT") 

# Set the max search time window (e.g., 90 minutes)
max_search_time <- 150

# Initialize a list to store results
travel_times_list_interact <- list()
latest_departure_times <- list()
latest_travel_times <- list()

# Simplify school data and join with student home tracts
sst_slim <- sst_new %>% 
  dplyr::select(name, school_number, am_bell, pm_bell, district_early_release_time) %>% 
  rename(name_sst = name)

# add bell times to student home tract dat
missing_student_home_tracts_with_bell_times <- every_missing_tract_school_combo %>% 
  left_join(sst_slim, by = c("prg_num" = "school_number")) %>% 
  # remove NACA Freedom and Democracy Academy II (which has closed )
  filter(name != "NACA Freedom and Democracy Academy II") %>% 
  left_join(school_info_join, by = c(
                                     "prg_num" = "prog_no")
  ) %>% 
  filter(!is.na(program)) %>% 
  mutate(id = row_number()) %>% 
  rename(prog_no = prg_num)


every_missing_tract_school_combo %>% 
  left_join(sst_slim, by = c("prg_num" = "school_number")) %>% 
  # remove NACA Freedom and Democracy Academy II (which has closed )
  filter(name != "NACA Freedom and Democracy Academy II") %>% 
  left_join(school_info_join, by = c(
                                     "prg_num" = "prog_no")
  ) %>% 
  filter(is.na(program)) %>% 
  mutate(id = row_number()) %>% 
  rename(prog_no = prg_num) %>% 
  group_by(name) %>% 
  count(sort = T)


# Test with a subset
sht_bell_test <- 
  missing_student_home_tracts_with_bell_times 


```

```{r}
for(i in 1:nrow(sht_bell_test )){
  
  # Get one row of data
  one_row <- sht_bell_test [i, ] %>% 
    st_drop_geometry() %>% 
    dplyr::select(-geometry) %>% 
    dplyr::select(-school_point,
                  -school_polygon)

  
  # Prepare data for r5r
  one_row_test <- one_row %>% 
    mutate(school_lat_lon = paste0(school_point_lat, ",", school_point_lon),
           home_tract_lat_lon = paste0(tract_point_lat, ",", tract_point_lon)) %>% 
    pivot_longer(
      cols = c("home_tract_lat_lon", "school_lat_lon"),
      names_to = "coord_type",
      values_to = "coord") %>% 
    mutate(id = ifelse(
      coord_type == "home_tract_lat_lon", 
      janitor::make_clean_names(census_tract), 
      str_replace_all(stringr::str_to_lower(program), " ", "_"))
      ) %>% 
    mutate(lat = map_dbl(.$coord, ~as.double(str_split(., ",")[[1]][1])),
           lon = map_dbl(.$coord, ~as.double(str_split(., ",")[[1]][2]))) %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
    mutate(geometry = st_make_valid(geometry)) %>% 
    # re-separate lat and lon
    mutate(lat = st_coordinates(geometry)[, 2],
           lon = st_coordinates(geometry)[, 1]) %>% 
    dplyr::select(prog_no,
                  # program,
                  name,
                  census_tract,
                  # students,
                  bldg_num,
                  grades,
                  type,
                  coord_type,
                  id,
                  lat,
                  lon,
                  am_bell,
                  pm_bell) %>%
    # five minutes before the bell will be the latest arrival time we'll allow 
    mutate(latest_arrival = as_hms((am_bell) - 5 * 60)) 
  
  # Initialize the latest valid departure time as the earliest possible
  latest_departure_time <- NA
  
  # Prepare data for joining with school data
  school_join_object <- one_row_test %>% 
    filter(coord_type != "home_tract_lat_lon") %>% 
    dplyr::select(id, 
                  prog_no, 
                  bldg_num, 
                  name, type, 
                  grades, 
                  # students, 
                  am_bell, 
                  pm_bell)
  
  # Loop to find the latest valid departure time
  for (j in seq(from = max_search_time, to = 10, by = -10)) {
    
    # start from the earliest possible time (in our case, it'll be 2 hours before the morning bell)
    departure_time <- as.POSIXct(paste(date, one_row_test[1,]$am_bell), format = "%Y-%m-%d %H:%M:%S") - (j * 60)
    
    # Calculate travel time using r5r
    travel_times <- travel_time_matrix(
      r5r_core = r5r_core,
      origins = one_row_test[one_row_test$coord_type == "home_tract_lat_lon", ],
      destinations = one_row_test[one_row_test$coord_type == "school_lat_lon", ],
      mode = mode,
      max_walk_time = 150,
      departure_datetime = departure_time,
      # verbose = TRUE
    ) %>% 
      mutate(departure_time = departure_time,
             on_time = travel_time_p50 <= as.numeric(difftime(as.POSIXct(paste(date, one_row_test[1,]$latest_arrival), format = "%Y-%m-%d %H:%M:%S"), departure_time, units = "mins")))
    
    # if travel_times is empty, return a departure time an hour before the am bell 
    
    if(nrow(travel_times) == 0) {
      departure_time_broken <- as.POSIXct(paste(date, one_row_test[1,]$am_bell), format = "%Y-%m-%d %H:%M:%S") - (60 * 60)
      
      latest_departure_time[[i]] <- departure_time_broken
    } else
    
    # Check if the student can arrive on time
    if(travel_times$on_time == TRUE & nrow(travel_times > 0)) {
      
      # if the arrival is on time, store the departure time and join the data with school data
      latest_departure_time <- departure_time
      
      travel_times_list[[i]] <-
        travel_times %>% 
        left_join(school_join_object, by = c("to_id" = "id"))
      
      
    } else {
      # If arrival time is too late, change the departure time to the next possible time in our sequence and start the loop again
      print(departure_time)
      # travel_times$departure_date <- latest_departure_time
      break
    }
  }

# Store the result
latest_departure_times[[i]] <- latest_departure_time

print(paste("Iteration", i, "complete.", "Latest departure time:", latest_departure_time))



gc()

}

# writing some placeholder code to handle the route combos that are stilll not working for us.
# in "latest_departure_times_quick_fix" we'll assume a departure time of 1 hour before the am bell
# "latest_departure_times_fucked" will be used to store the rows that are causing us trouble

# fucked_up_iterations %>% 
#   write_csv("data/output/fucked_up_iterations_01_14.csv")
# 
# latest_departure_times %>% 
#   write_rds("data/setup/latest_departure_times_not_real_combos.rds")
# 
# travel_times_list %>% 
#   write_rds("data/setup/travel_times_list_not_real_combos.rds")


travel_times_from_student_home_tracts <- 
  travel_times_list %>% 
  bind_rows()  %>% 
  mutate(arrival_time = departure_time + travel_time_p50 * 60) 

departure_times_
  

# latest_departure_times_fucked <- 
#   latest_departure_times %>% 
#   lapply(
#     # when an iteration has a length greater than 1, return an NA
#     function(x) if(length(x) > 1) NA else x
#   )
# 



latest_leaving_times_for_detailed_routing <- 
  travel_times_from_student_home_tracts %>% 
  mutate(prog_no = as.character(prog_no),
         from_id = str_remove_all(from_id, "x")
         # # convert arrival and departure times to EST
         # arrival_time = as.POSIXct(arrival_time, tz = "UTC") - hours(4),
         # departure_time = as.POSIXct(departure_time, tz = "UTC") - hours(4)
         ) %>% 
  left_join(school_info_join, 
            by = c(
              "prog_no" = "prog_no"
            ))  %>% 
  left_join(missing_student_home_tracts_with_bell_times,
            by = c(
              "from_id" = "census_tract",
              "prog_no" = "prog_no",
              "grades" = "grades",
              "name" = "name",
              "type" = "type",
              "bldg_num" = "bldg_num",
              "program" = "program",
              "am_bell" = "am_bell",
              "pm_bell" = "pm_bell"
            ))  %>%
    rename(census_tract = from_id) %>% 
  mutate(id = row_number()) %>% 
    dplyr::select(prog_no,
                  departure_time,
                  arrival_time,
                  program,
                  name,
                  census_tract,
                  # students,
                  bldg_num,
                  grades,
                  type,
                  # coord_type,
                  # id,
                  # lat,
                  # lon,
                  am_bell,
                  pm_bell,
                  school_point_lat,
                  school_point_lon,
                  tract_point_lat,
                  tract_point_lon,
                  geometry.x) %>% 
  rename(geometry = geometry.x) 

latest_leaving_times_for_detailed_routing %>% 
  write_rds("data/setup/latest_leaving_times_for_detailed_routing_not_real.rds")
```

```{r}
# Define modes of transit that we'll give our commuters access to 
mode <- c("WALK", "TRANSIT") 
i = 1


for(i in 1:nrow(latest_leaving_times_for_detailed_routing)){
  
  one_row <- latest_leaving_times_for_detailed_routing[i,]
  
  one_row_test <-
  one_row %>% 
    mutate(school_lat_lon = paste0(school_point_lat, ",", school_point_lon),
           home_tract_lat_lon = paste0(tract_point_lat, ",", tract_point_lon)) %>% 
    pivot_longer(
      cols = c("home_tract_lat_lon", "school_lat_lon"),
      names_to = "coord_type",
      values_to = "coord") %>% 
    mutate(id = ifelse(
      coord_type == "home_tract_lat_lon", 
      janitor::make_clean_names(census_tract), 
      str_replace_all(stringr::str_to_lower(program), " ", "_"))
      ) %>% 
    mutate(lat = map_dbl(.$coord, ~as.double(str_split(., ",")[[1]][1])),
           lon = map_dbl(.$coord, ~as.double(str_split(., ",")[[1]][2]))) %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
    mutate(geometry = st_make_valid(geometry)) %>% 
    # re-separate lat and lon
    mutate(lat = st_coordinates(geometry)[, 2],
           lon = st_coordinates(geometry)[, 1]) %>%
    dplyr::select(prog_no,
                  program,
                  name,
                  census_tract,
                  # students,
                  bldg_num,
                  grades,
                  type,
                  coord_type,
                  id,
                  lat,
                  lon,
                  am_bell,
                  pm_bell)
  
  # make detailed routes directory inside of today's poa folder
  dir.create(here::here("data/poa", today, "detailed-routes_not_real"), showWarnings = F)

  df <- 
    detailed_itineraries(
      r5r_core,
      origins = one_row_test[1,],
      destinations = one_row_test[2,],
      mode = mode,
      departure_datetime = one_row$departure_time,
      shortest_path = T,
      output_dir = here::here("data/poa", today, "detailed-routes_not_real")
      )
  
  gc()
  
  Sys.sleep(0.1)
  
}

```

# Read each csv file into a list of dataframes

```{r}
# read in the csv files

file_path <- here::here("data/poa", today, "detailed-routes_not_real")

file_paths <- list.files(here::here(file_path), full.names = T) %>% unique

df_list <- list()

df_list <- lapply(file_paths, read_csv)
```

```{r}
school_list <-
  df_list %>% 
  # mutate all from_id and to_id to character
  map(~mutate(.x, from_id = as.character(from_id), to_id = as.character(to_id))) %>%
  map(~mutate(.x, to_id = map_chr(to_id, ~str_remove_all(.x, "stop:")))) %>% 
  map(~mutate(.x, to_lat = as.character(to_lat), to_lon = as.character(to_lat))) %>% 
  map(~mutate(.x, departure_time = as.character(departure_time))) %>%
  map(~mutate(.x, total_distance = as.numeric(total_distance))) %>% 
  map(~mutate(.x, mode = as.character(mode))) %>% 
  map(~mutate(.x, wait = as.numeric(wait))) %>%
  bind_rows() %>% 
  group_by(from_id, to_id) %>%
  mutate(route = row_number()) 

geom <- school_list %>% 
  head(2) %>% 
  .$geometry

school_list %>% 
  st_as_sf(sf_column_name = "geometry")

school_list$geometry <- st_as_sfc(school_list$geometry)

school_list %>% 
  st_as_sf(crs = 4326) %>% 
  st_write(here::here("data/poa", today, "every_leg_every_trip.shp"))

school_list %>% 
  # filter(from_id == "x24510280302",
  #        to_id == "mer-vo_h") %>% 
  st_as_sf(crs = 4326) %>% 
  mapview(zcol = "mode") +
  mapview(baltimore_grid_sampled) +
  mapview(bmore_tracts_no_parks, alpha.regions = 0.2) +
  # add school shapes 
  mapview(hs_parcels_buffer, col.regions = "red") 

```

