---
title: "School Trip Routing (Final)"
format: html
editor: visual
---

# Testing r5r

## Setup

### Load packages

```{r}
library(devtools)
library(elevatr)
library(r5r)
library(sf)
library(osmextract)
library(rJava)
library(tidyverse)
library(geobr)
library(here)
library(tigris)
library(tidycensus)
library(mapview)
library(RColorBrewer)
library(readxl)
library(janitor)
library(httr)
library(hms)
library(raster)
library(terra)
library(osmdata)

# turn off scientific notation
options(scipen = 999)

```

```{r}

pbf_file <- "/Users/greg.morton/Documents/baltimoreBanner/Projects/2024/high_school_transportation_routing/data/poa/2025-01-17/baltimore.osm.pbf"



  # Read the PBF file using sf
  streets <- sf::st_read(pbf_file, layer = "lines", quiet = TRUE)

  # Filter for roads (you can modify these conditions based on your needs)
  roads <- streets[streets$highway %in% c("motorway", "trunk", "primary", "secondary",
                                        "tertiary", "residential", "unclassified"), ]
#   
  # Create an interactive visualization
  map <- mapview(roads,
                zcol = "highway",  # Color by road type
                col.regions = rainbow(length(unique(roads$highway))),
                layer.name = "Street Network",
                map.types = c("OpenStreetMap", "Esri.WorldImagery"),
                legend = TRUE,
                popup = TRUE)
  
  map



mizod <- st_read("/Users/greg.morton/Documents/baltimoreBanner/Projects/2024/high_school_transportation_routing/data/shapes/MIZOD/MIZOD.shp") %>% 
  st_union() %>% 
  st_transform(4326) 


```

### Load school location data

```{r}
# school points 
D_school_locs <- st_read("data/shapes/city_schools/Baltimore_City_Schools.shp") %>% 
  st_transform(4326) %>% 
  st_set_crs(4326)


# filter points down to high schools and middle schools 
hs_locs_join <- 
  D_school_locs %>% 
  filter(str_detect(class, "9 - 12|6 - 12|6 - 8|5 - 8")) %>% 
  mutate(join_addr = str_to_lower(address), 
  ) 

# df version for joining
hs_locs_join_df <- 
  hs_locs_join %>% 
  as.data.frame() %>% 
  mutate(prg_num = as.character(prg_num))

# read in school parcels 
D_school_parcels <- st_read("data/shapes/school_parcels/school_parcels.geojson") %>% 
  distinct(.keep_all = T) %>%
  mutate(schools_at_parcel = 
           case_when(
             schools_at_parcel == "Crossroads School, The" ~ "The Crossroads School",
             schools_at_parcel == "Belair-Edison School, The" ~ "The Belair-Edison School",
             schools_at_parcel == "Dallas F. Nicholas, Sr., Elementary School" ~ "Dallas F. Nicholas Sr. Elementary School",
             schools_at_parcel == "Green School Of Baltimore, The" ~ "The Green School Of Baltimore",
             schools_at_parcel == "Reach! Partnership School, The" ~ "The Reach! Partnership School",
             schools_at_parcel == "Dr. Bernard Harris, Sr., Elementary School" ~ "Dr. Bernard Harris Sr. Elementary School",
             schools_at_parcel == "Historic Samuel Coleridge-Taylor Elementary School, The, Joseph C. Briscoe Academy" ~ "The Historic Samuel Coleridge-Taylor Elementary School, The Joseph C. Briscoe Academy",
             TRUE ~ schools_at_parcel
           )) %>% 
  st_transform(4326) %>% 
  st_set_crs(4326)

# D_school_parcels %>% 
#   st_write("/Users/greg.morton/Documents/baltimoreBanner/Projects/2024/high_school_transportation_routing/data/shapes/school_parcels/parcels_01_22.geojson")




# join school parcels to high school locations
hs_parcels_buffer <- 
  D_school_parcels %>% 
  st_join(hs_locs_join) %>%
  dplyr::select(school_name, name, join_addr, fulladdr) %>% 
  filter(!is.na(name)) %>% 
  distinct(name, .keep_all = T) %>% 
  as.data.frame() %>% 
  left_join(hs_locs_join_df,
            by = c(name = "name")) %>% 
  filter(str_detect(class, "9 - 12|6 - 12|6 - 8|5 - 8")) %>% 
  st_as_sf() %>%
  st_transform(29902)


# join df for hs parcels (we just need school number from here)
hs_parcel_join <- 
  hs_parcels_buffer %>% 
  as.data.frame() %>% 
  mutate(school_number = as.character(prg_num)) %>% 
  ungroup()

# get start stop times 
start_stop_times <- read_csv("data/setup/start_stop_times.csv") %>% 
  janitor::clean_names() %>% 
  mutate(school_number = as.character(school_number))

# select relevant columns 
sst_new <- start_stop_times %>% 
  left_join(hs_parcel_join, by = c("school_number" = "school_number")) %>% 
  filter(!is.na(name)) %>% 
  dplyr::select(school_number, name, am_bell, pm_bell, district_early_release_time, address.y
                ) 

# load functions I need to access MTA API
source("functions/mta_api_functions.R")


```

### Load Baltimore City Shape

```{r}
# get baltimmore city census tracts 
bmore_tracts <- tigris::tracts(
  state = "MD", 
  county = "Baltimore City"
) %>%
  tigris::erase_water() %>%
  st_transform(4326) %>%
  # First make valid
  st_make_valid() %>%
  # Remove any empty geometries
  filter(!st_is_empty(.)) %>%
  # Extract polygons
  st_collection_extract("POLYGON") %>%
  # Remove any potential duplicate vertices and fix other issues
  st_buffer(0) %>%
  # Group and union
  group_by(GEOID, NAME) %>% 
  summarise(geometry = st_union(geometry), .groups = "drop")

# add parks so we can remove them 
parks <- st_read("data/shapes/Parks/Parks.shp") %>% st_transform(32618) 

# add a slight buffer to the parks so we can remove them from the tracts
parks <- st_buffer(parks, 50) %>% st_union() %>% st_transform(4326)

# add schools so we can remove them 
schools <- hs_parcels_buffer %>% 
  st_buffer(100) %>% 
  st_union() %>% 
  st_transform(4326)


# load EU Global Human Settlement Layer Raster
grid <- raster("data/raw/GHS_BUILT_S_E2030_GLOBE_R2023A_54009_100_V1_0_R5_C12/GHS_BUILT_S_E2030_GLOBE_R2023A_54009_100_V1_0_R5_C12.tif")

#Turn it into points
grid <- projectRaster(grid, crs = st_crs(4326)$proj4string)

# Make a bounding box that we will use for cropping the raster
bbox <- extent(st_bbox(bmore_tracts))

# another bounding box that we'll use to crop the OSM.PBF file later
area_bbox <- st_bbox(bmore_tracts) %>% 
  st_as_sfc() %>% 
  st_transform(4326) %>% 
  st_buffer(3000) %>% 
  st_bbox() %>% 
  st_as_sfc() %>% 
  st_transform(4326) %>% 
  st_bbox()
```

```{r}
# cropped_raster <- crop(grid, bbox)
# 
# baltimore_grid <- rasterToPoints(cropped_raster, spatial = TRUE)

# get Baltimore elevation data from elevatr
# Convert bbox to sf object for get_elev_raster
aoi <- st_as_sfc(st_bbox(area_bbox)) %>%
  st_sf() %>%
  st_set_crs(4326)

# Step 2: Get elevation data using elevatr
# zoom = 9-14 determines resolution (higher = more detailed but larger file)
elevation_raster <- get_elev_raster(locations = aoi, z = 12, clip = "locations")

elevation_terra <- rast(elevation_raster)  # Convert to terra format

writeRaster(elevation_terra, 
            filename = "data/poa/elevation.tif",
            overwrite = TRUE)

```

```{r}
# version of tracts with no parks 
bmore_tracts_no_parks <- 
  bmore_tracts %>% 
  st_difference(parks) %>%
  st_difference(schools) %>%
  st_difference(mizod) 


# # get a new grid sample
# baltimore_grid_sampled <-
#   baltimore_grid %>%
#   as.data.frame() %>%
#   st_as_sf(coords = c("x", "y"), crs = 4326) %>%
#   st_difference(parks) %>%
#   st_difference(schools) %>%
#   st_difference(mizod) %>%
#   st_join(bmore_tracts) %>%
#   filter(!is.na(GEOID)) %>%
#   # filter(!is.na(median_income)) %>%
#   group_by(GEOID) %>%
#   filter(GHS_BUILT_S_E2030_GLOBE_R2023A_54009_100_V1_0_R5_C12 > 3000) %>%
#   # Calculate points_to_sample based on population proportion
#   # Use group_modify to sample points
#   # group_modify(~ slice_sample(.x, n = first(.x$points_to_sample), replace = TRUE)) %>%
#   # sample one point from each tract
#   group_modify(~ slice_sample(.x, n = 1, replace = TRUE)) %>%
#   ungroup() %>%
#   st_as_sf(crs = 4326) %>%
#   st_make_valid() %>%
#   mutate(geometry = st_make_valid(geometry)) %>%
#   mutate(lon = map_dbl(geometry, ~st_coordinates(.)[1]),
#          lat = map_dbl(geometry, ~st_coordinates(.)[2])) %>%
#   ungroup() %>%
#   mutate(id = row_number())
# # 
# baltimore_grid_sampled <-
#   st_read("data/shapes/bmore_grid_sampled_01_17.geojson") %>%
#   st_as_sf()
#   



baltimore_grid_sampled <-
  st_read("data/shapes/bmore_grid_sampled_01_27.geojson") %>%
  st_as_sf()


# broken_start <-
#   sht_bell_test %>%
#   .[4140,] %>%
#   dplyr::select(geometry) %>%
#   st_as_sf()
# 
# baltimore_grid_sampled %>%
#   mapview(zcol = "GHS_BUILT_S_E2030_GLOBE_R2023A_54009_100_V1_0_R5_C12", legend = F) +
# mapview(bmore_tracts_no_parks) 





```

```{r}
  # mapview(hs_parcels_buffer, col.regions = "red", legend = F) +
  # # mapview(broken_start, col.regions = "yellow", legend = F) +
  # map +
  # mapview(bmore_tracts, zcol = "GEOID", legend = F) +
  # mapview(fucked_up_iterations_to, zcol = "row_number") +
  # mapview(fucked_up_iterations_from, zcol = "row_number")



  
  
  
# 






```

```{r}
# make a join version with lat and lon for r5r
bmore_tracts_join <-
  bmore_tracts %>% 
  clean_names() %>% 
  dplyr::select(geoid, name, geometry) %>% 
  rename(tract_name = name) %>%
  st_as_sf() %>% 
  st_transform(4326) %>% 
  st_difference(parks) %>% 
  st_make_valid() %>%
  mutate(
    valid_geometry = st_make_valid(geometry),  # Ensure geometries are valid
    tract_point = st_centroid(valid_geometry),  # Calculate the centroid of the land area,
    tract_point_lat = st_coordinates(tract_point)[, 2],  # Extract the latitude of the centroid
    tract_point_lon = st_coordinates(tract_point)[, 1]  # Extract the longitude of the centroid
  ) 


```

### Load student location data

The 'student_home_tract_XX' objects hold info about where a student is going and where they are coming from.

There is a row for each combination of home census tract and school in the data.

```{r}
# read student home tract data 
student_home_tracts_23 <- 
  read_xlsx("data/home_tracts/REVISED Banner Request-Enroll by Prog and Tract-2024.08.08.xlsx", sheet = 2, skip = 5) %>% 
  clean_names() %>% 
  mutate(prog_no = as.character(prog_no)) %>%
  # change all colnames to character
  mutate(across(-c(prog_no, program, enrollment), ~ as.numeric(as.character(.)))) %>%
  # Pivot longer to convert census tract columns into rows
  pivot_longer(
    cols = -c(prog_no, program, enrollment),
    names_to = "census_tract",
    values_to = "students"
  ) %>% 
  left_join(hs_parcel_join, by = c("prog_no" = "school_number")) %>%
  dplyr::select(prog_no, program, enrollment, school_name, name, everything()) %>%
  mutate(students = ifelse(is.na(students), 5, students)) %>%
  filter(!is.na(school_name),
         students > 0
         )  %>%
  dplyr::select(-x, -y, -join_addr.y, -GlobalID, -class, -category, -OBJECTID, -join_addr.x) %>% 
  rename(school_polygon = geometry.x,
         school_point = geometry.y) %>% 
  mutate(school_point = st_make_valid(school_point)) %>% 
  mutate(school_point_lat = st_coordinates(school_point)[, 2],
         school_point_lon = st_coordinates(school_point)[, 1],
         census_tract = str_remove_all(census_tract, "x")) %>% 
  as.data.frame() %>%
  # There are two ways that we can assign students home tracts
  # first we can simply use "bmore_tracts_join", an sf object that simply takes the centerpoint of each tract
  # the upside here is that it's easier and its consistent
  # The downside is that sometimes it can put our point at a place we know no one lives, like a park or industrial area
  # left_join(bmore_tracts_join, by = c("census_tract" = "geoid")) 
  # The second method is to sample human settlement points from the raster grid and join them to the tracts
  # We'll then take a random point from each tract to represent the students home location
  # This is probably more accurate but it's also more computationally expensive
  # We'll do both and compare the results
  left_join(baltimore_grid_sampled, by = c("census_tract" = "GEOID")) %>% 
  rename(
    tract_point_lat = lat,
    tract_point_lon = lon
  ) 


# make enrollment df
enrollment_23 <- 
  read_xlsx("data/home_tracts/REVISED Banner Request-Enroll by Prog and Tract-2024.08.08.xlsx", sheet = 2, skip = 5) %>% 
  clean_names() %>% 
  mutate(prog_no = as.character(prog_no)) %>%
  # change all colnames to character
  mutate(across(-c(prog_no, program, enrollment), ~ as.numeric(as.character(.)))) %>%
  # Pivot longer to convert census tract columns into rows
  pivot_longer(
    cols = -c(prog_no, program, enrollment),
    names_to = "census_tract",
    values_to = "students"
  ) %>% 
  # left_join(hs_parcel_join, by = c("prog_no" = "school_number")) %>%
  dplyr::select(prog_no, program, enrollment, everything()) %>% 
  mutate(students = ifelse(is.na(students), 5, students)) %>%
  filter(!is.na(program),
         students > 0
         )  %>% 
  dplyr::select(
    prog_no,
    program,
    enrollment
  ) %>% 
  mutate(program_join = str_to_lower(str_replace_all(program, " ", "_"))) %>% 
  group_by(program) %>% 
  distinct() %>% 
  ungroup() %>% 
  left_join(hs_locs_join_df, by = c("prog_no" = "prg_num")) %>% 
  filter(!is.na(name))


# get enrollment by tract 
enrollment_by_tract <-
  student_home_tracts_23 %>%
  dplyr::select(
    census_tract,
    name,
    students,
    zipcode
  )

  
  

```

### Load MTA stuff

```{r}
# mta routes 
routes_url <- "https://api.goswift.ly/info/mta-maryland/routes"

# light rail url for api call
light_rail_url <- "https://api.goswift.ly/info/mta-maryland-light-rail/routes/"

# metro url for api call
metro_url <- "https://api.goswift.ly/info/mta-maryland-metro/routes/"

# get bus api response
routes_response <- VERB("GET", routes_url, add_headers('Authorization' = '0352517c0285694758568f98b6f013ba'), content_type("application/octet-stream"), accept("application/json"))

# light rail api response
light_rail_response <- VERB("GET", light_rail_url, add_headers('Authorization' = '0352517c0285694758568f98b6f013ba'), content_type("application/octet-stream"), accept("application/json"))

# metro api response 
metro_response <- VERB("GET", metro_url, add_headers('Authorization' = '0352517c0285694758568f98b6f013ba'), content_type("application/octet-stream"), accept("application/json"))

# routes json
routes_json <- content(routes_response, "text")

# lr json
light_rail_json <- content(light_rail_response, "text")

# metro json
metro_json <- content(metro_response, "text")

# turn em all into dataframes
mta_routes <- 
  jsonlite::fromJSON(routes_json, simplifyDataFrame = T, flatten = F) %>% 
  .$data %>% 
  .$routes %>% 
  clean_names

light_rail_routes <-
    jsonlite::fromJSON(light_rail_json, simplifyDataFrame = T, flatten = F) %>% 
  .$data %>% 
  .$routes %>% 
  clean_names

metro_routes <-
    jsonlite::fromJSON(metro_json, simplifyDataFrame = T, flatten = F) %>%
  .$data %>%
  .$routes %>%
  clean_names

all_routes <- 
  bind_rows(mta_routes, light_rail_routes, metro_routes) %>% 
  mutate(short_name = str_remove_all(short_name, " "))

D_school_parcels <- st_read("data/shapes/school_parcels/school_parcels.geojson") %>% 
  distinct(.keep_all = T) %>%
  mutate(schools_at_parcel = 
           case_when(
             schools_at_parcel == "Crossroads School, The" ~ "The Crossroads School",
             schools_at_parcel == "Belair-Edison School, The" ~ "The Belair-Edison School",
             schools_at_parcel == "Dallas F. Nicholas, Sr., Elementary School" ~ "Dallas F. Nicholas Sr. Elementary School",
             schools_at_parcel == "Green School Of Baltimore, The" ~ "The Green School Of Baltimore",
             schools_at_parcel == "Reach! Partnership School, The" ~ "The Reach! Partnership School",
             schools_at_parcel == "Dr. Bernard Harris, Sr., Elementary School" ~ "Dr. Bernard Harris Sr. Elementary School",
             schools_at_parcel == "Historic Samuel Coleridge-Taylor Elementary School, The, Joseph C. Briscoe Academy" ~ "The Historic Samuel Coleridge-Taylor Elementary School, The Joseph C. Briscoe Academy",
             TRUE ~ schools_at_parcel
           )) %>% 
  st_transform(4326) %>% 
  st_set_crs(4326) %>% 
  mutate(school_number = as.character(school_number)) %>% 
  left_join(enrollment_23, by = c("school_number" = "prog_no")) %>% 
  filter(!is.na(program_join)) 
```

### Setup r5r

```{r}
# start rjava and allocate enough memory for r5r
rJava::.jinit()
options(java.parameters = "-Xmx50g")
rJava::.jcall("java.lang.System", "S", "getProperty", "java.version")
# un-comment for session details 
# session_info()
# r5r::r5r_sitrep()

# load baltimore shape un-comment if you need to load osm file
# baltimore_shape <-
#   tigris::places("Baltimore", state = "MD", class = "city") %>%
#   filter(NAME == "Baltimore") %>%
#   st_as_sf()  %>%
#   st_transform(4326) 

# define the modes of transit we need gtfs files for
modes <- c("light-rail", "local-bus", "metro")

# get todays date in the format we need
today <- lubridate::today() %>% 
  as.character() %>%
  str_split(" ") %>% 
  map_chr(1)

# create a directory for the data
dir.create(here::here("data/poa/", today), showWarnings = F)


# grab today's GTFS files 
lapply(modes, function(x) {
  gtfs_url <- paste0("https://feeds.mta.maryland.gov/gtfs/", x)
  download.file(gtfs_url, paste0("data/poa/",today, "/", x, "_gtfs.zip"))
  })

# grab .tif elevation 
# tst <- st_read("/Users/greg.morton/Documents/baltimoreBanner/Projects/2024/high_school_transportation/data/NED10M_SpatialMetadata/NED10M_fe2947_October2018.shp") %>% 
#   st_transform(4326) %>% 
#   st_set_crs(4326) 

# get osm data for baltimore
# oe_get("us/maryland", stringsAsFactors = FALSE, quiet = FALSE)
# oe_download_directory()

# get the bounding box of the baltimore shape
# area_bbox <- sf::st_bbox(baltimore_shape)

# path to osmosis
osmosis_path <- "/opt/homebrew/bin/osmosis"

# path to the large .pbf file
large_pbf_path <- "data/poa/maryland-latest.osm.pbf"


# path where you want to save the smaller .pbf file
smaller_pbf <-"data/poa/baltimore_complete_ways_complete_relations.osm.pbf"

# prepare call to osmosis
osmosis_cmd <- sprintf(
  "%s --read-pbf file=%s --bounding-box completeWays=yes left=%s bottom=%s right=%s top=%s --write-pbf file=%s",
  osmosis_path, large_pbf_path,
  area_bbox["xmin"], area_bbox["ymin"], area_bbox["xmax"], area_bbox["ymax"],
  smaller_pbf
)

# call to osmosis
system(osmosis_cmd, intern = TRUE)

# copy the OSM PBF files from data/osm_pbf into today's poa directory
path = paste0("data/poa/", as.character(today), "/baltimore.osm.pbf")

file.copy(
  from = "data/poa/baltimore_complete_ways_complete_relations.osm.pbf",
  to = paste0("data/poa/", as.character(today), "/baltimore.osm.pbf"),
  overwrite = TRUE
)

# copy .tif file 
file.copy(
  from = "data/poa/elevation.tif",
  to = paste0("data/poa/", as.character(today), "/elevation.tif"),
  overwrite = TRUE
)



# setup r5r
r5r_core <- setup_r5(here::here(paste0("data/poa/", "2025-02-03")),
                     temp_dir = F,
                     verbose = T,
                     overwrite = F
                     )

r5r_core

# stop r5r
r5r::stop_r5(r5r_core)
# rJava::.jgc(R.gc = TRUE)
```

## Part I: Determine the latest time students can leave home and still get to school on time

Before we start modeling which routes students take to school, we need to determine how late they can expect to leave and still make it to school on time. We cannot run the 'detailed_iternaries' function without a departure time.

To do this, we'll use the 'time_travel_matrix' function from r5r to calculate the travel time from each student's home to their school. We'll loop through a range of departure times and find the latest time that still allows the student to arrive at school on time.

```{r}
# Define modes of transit
mode <- c("WALK", "TRANSIT") 

# Set the max search time window (e.g., 90 minutes)
max_search_time <- 150

# Initialize a list to store results
travel_times_list <- list()
latest_departure_times <- list()
latest_travel_times <- list()

# Simplify school data and join with student home tracts
sst_slim <- sst_new %>% 
  dplyr::select(name, school_number, am_bell, pm_bell, district_early_release_time) %>% 
  rename(name_sst = name)

# add bell times to student home tract dat
student_home_tracts_with_bell_times <- student_home_tracts_23 %>% 
  left_join(sst_slim, by = c("prog_no" = "school_number")) 

# hard coded date
date <- today

# Test with a subset
sht_bell_test <- 
  student_home_tracts_with_bell_times 


# weird_tract <- 
#   sht_bell_test %>% 
#   mutate(row_number = as.character(row_number())) %>% 
#   filter(census_tract == "24510250500") 
# 
# fucked_up_rows <- 
#   c("1056", "1783", "2358", "2945", "4140")
# 
# fucked_up_iterations <-
#   weird_tract %>% 
#   filter(row_number %in% fucked_up_rows) %>% 
#   mutate(geometry = st_make_valid(geometry),
#          school_point = st_make_valid(school_point)) 
# 
# fucked_up_iterations_to <- 
#   fucked_up_iterations %>% 
#   dplyr::select(school_point, row_number) %>% 
#   st_as_sf()
# 
# fucked_up_iterations_from <- 
#   fucked_up_iterations %>% 
#   dplyr::select(geometry, row_number) %>% 
#   st_as_sf()
# 
# 
#    
# date
```

```{r}
for(i in 1:nrow(sht_bell_test )){
  
  # Get one row of data
  one_row <- sht_bell_test [i, ] %>% 
    st_drop_geometry() %>% 
    dplyr::select(-geometry) %>% 
    dplyr::select(-school_point,
                  -school_polygon)

  
  # Prepare data for r5r
  one_row_test <- one_row %>% 
    mutate(school_lat_lon = paste0(school_point_lat, ",", school_point_lon),
           home_tract_lat_lon = paste0(tract_point_lat, ",", tract_point_lon)) %>% 
    pivot_longer(
      cols = c("home_tract_lat_lon", "school_lat_lon"),
      names_to = "coord_type",
      values_to = "coord") %>% 
    mutate(id = ifelse(
      coord_type == "home_tract_lat_lon", 
      janitor::make_clean_names(census_tract), 
      str_replace_all(stringr::str_to_lower(program), " ", "_"))
      ) %>% 
    mutate(lat = map_dbl(.$coord, ~as.double(str_split(., ",")[[1]][1])),
           lon = map_dbl(.$coord, ~as.double(str_split(., ",")[[1]][2]))) %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
    mutate(geometry = st_make_valid(geometry)) %>% 
    # re-separate lat and lon
    mutate(lat = st_coordinates(geometry)[, 2],
           lon = st_coordinates(geometry)[, 1]) %>%
    dplyr::select(prog_no,
                  program,
                  name,
                  census_tract,
                  students,
                  bldg_num,
                  grades,
                  type,
                  coord_type,
                  id,
                  lat,
                  lon,
                  am_bell,
                  pm_bell) %>%
    # five minutes before the bell will be the latest arrival time we'll allow 
    mutate(latest_arrival = as_hms((am_bell) - 5 * 60)) 
  
  # Initialize the latest valid departure time as the earliest possible
  latest_departure_time <- NA
  
  # Prepare data for joining with school data
  school_join_object <- one_row_test %>% 
    filter(coord_type != "home_tract_lat_lon") %>% 
    dplyr::select(id, prog_no, bldg_num, name, type, grades, students, am_bell, pm_bell)
  
  # Loop to find the latest valid departure time
  for (j in seq(from = max_search_time, to = 10, by = -10)) {
    
    # start from the earliest possible time (in our case, it'll be 2 hours before the morning bell)
    departure_time <- as.POSIXct(paste(date, one_row_test[1,]$am_bell), format = "%Y-%m-%d %H:%M:%S") - (j * 60)
    
    # Calculate travel time using r5r
    travel_times <- travel_time_matrix(
      r5r_core = r5r_core,
      origins = one_row_test[one_row_test$coord_type == "home_tract_lat_lon", ],
      destinations = one_row_test[one_row_test$coord_type == "school_lat_lon", ],
      mode = mode,
      max_walk_time = 150,
      departure_datetime = departure_time,
      # verbose = TRUE
    ) %>% 
      mutate(departure_time = departure_time,
             on_time = travel_time_p50 <= as.numeric(difftime(as.POSIXct(paste(date, one_row_test[1,]$latest_arrival), format = "%Y-%m-%d %H:%M:%S"), departure_time, units = "mins")))
    
    # if travel_times is empty, return a departure time an hour before the am bell 
    
    if(nrow(travel_times) == 0) {
      departure_time_broken <- as.POSIXct(paste(date, one_row_test[1,]$am_bell), format = "%Y-%m-%d %H:%M:%S") - (60 * 60)
      
      latest_departure_time[[i]] <- departure_time_broken
    } else
    
    # Check if the student can arrive on time
    if(travel_times$on_time == TRUE & nrow(travel_times > 0)) {
      
      # if the arrival is on time, store the departure time and join the data with school data
      latest_departure_time <- departure_time
      
      travel_times_list[[i]] <-
        travel_times %>% 
        left_join(school_join_object, by = c("to_id" = "id"))
      
      
    } else {
      # If arrival time is too late, change the departure time to the next possible time in our sequence and start the loop again
      print(departure_time)
      # travel_times$departure_date <- latest_departure_time
      break
    }
  }

# Store the result
latest_departure_times[[i]] <- latest_departure_time

print(paste("Iteration", i, "complete.", "Latest departure time:", latest_departure_time))



gc()

}

# writing some placeholder code to handle the route combos that are stilll not working for us.
# in "latest_departure_times_quick_fix" we'll assume a departure time of 1 hour before the am bell
# "latest_departure_times_fucked" will be used to store the rows that are causing us trouble

# fucked_up_iterations %>% 
#   write_csv("data/output/fucked_up_iterations_01_14.csv")

latest_departure_times_fucked <- 
  latest_departure_times %>% 
  lapply(
    # when an iteration has a length greater than 1, return an NA
    function(x) if(length(x) > 1) NA else x
  )

latest_departure_times_fucked %>% 
  write_rds("data/setup/latest_departure_times_fucked_01_17.rds")

latest_departure_times_quick_fix <-
  latest_departure_times %>% 
  lapply(
    # when an iteration has a length greater than 1, print the first element 
    function(x) if(length(x) > 1) print(x[1]) else x
  )

latest_departure_times_quick_fix %>% 
    write_rds("data/setup/latest_departure_times_quick_fix_01_17.rds")
```

```{r}
travel_times_from_student_home_tracts <- 
  travel_times_list %>% 
  bind_rows()  %>% 
  mutate(arrival_time = departure_time + travel_time_p50 * 60) 

# here's our save point
# write_rds(travel_times_from_student_home_tracts, paste0("data/poa/", today, "/travel_times_df.rds"))

read_rds(paste0("data/poa/", today, "/travel_times_df.rds")) %>% 
  mutate(arrival_time = departure_time + travel_time_p50 * 60) %>% 
  write_csv("data/output/travel_times_from_student_home_tracts_01_17.csv")


travel_times_from_student_home_tracts %>% 
  summarise(mean_travel_time = weighted.mean(travel_time_p50, students, na.rm = T),
            median_travel_time = median(travel_time_p50, students, na.rm = T),
            sd_travel_time = sd(travel_time_p50, na.rm = T),
            min_travel_time = min(travel_time_p50),
            max_travel_time = max(travel_time_p50)) 



```

### What if you have a car?

We'll model latest departure time in the same way here. Since modeling transit by car isn't as complicated, it's a little less computationally expensive.

We'll also have to make a decision here on whether to even run detailed routing for car trips.

```{r}
# Define modes of transit
# mode <- c("CAR") 
# 
# # Set the max search time window (e.g., 90 minutes)
# max_search_time <- 120
# 
# # Initialize a list to store results
# travel_times_list <- list()
# latest_departure_times <- list()
# latest_travel_times <- list()
# 
# # Simplify school data and join with student home tracts
# sst_slim <- sst_new %>% 
#   dplyr::select(name, school_number, am_bell, pm_bell, district_early_release_time) %>% 
#   rename(name_sst = name)
# 
# student_home_tracts_with_bell_times <- student_home_tracts_23 %>% 
#   left_join(sst_slim, by = c("prog_no" = "school_number")) 
# 
# # Get the current date for use in departure times
# date <- lubridate::now() %>% 
#   as.character() %>%
#   str_split(" ") %>% 
#   map_chr(1)
# 
# # Test with a subset
# sht_bell_test <- 
#   student_home_tracts_with_bell_times %>% 
#   mutate(join_program = str_to_lower(str_replace_all(program, " ", "_")),
#          census_tract = as.character(census_tract)) %>%
#   left_join(departure_times_10_02,
#             by = c("join_program" = "to_id",
#               "census_tract" = "from_id")
#             ) %>% 
#   # convert departure time to unix timestamp 
#   mutate( 
#     id = row_number(),
#     departure_time_unix = as.numeric(departure_time),
#     departure_hour = hour(departure_time),
#     departure_minute = minute(departure_time),
#     row_number = row_number()
#     ) 
# 
# 
# sht_bell_test %>%
#   write_csv("data/output/sht_bell_test.csv")
# 
# # Loop through each row in sht_bell_test
# for(i in 1:nrow(sht_bell_test)){
#   
#   one_row <- sht_bell_test[i,]
#   
#   # Prepare data for r5r
#   one_row_test <- one_row %>% 
#     mutate(school_lat_lon = paste0(school_point_lat, ",", school_point_lon),
#            home_tract_lat_lon = paste0(tract_point_lat, ",", tract_point_lon)) %>%
#     pivot_longer(
#       cols = c("home_tract_lat_lon", "school_lat_lon"),
#       names_to = "coord_type",
#       values_to = "coord") %>% 
#     mutate(id = ifelse(
#       coord_type == "home_tract_lat_lon", 
#       janitor::make_clean_names(census_tract), 
#       str_replace_all(stringr::str_to_lower(program), " ", "_"))
#       ) %>% 
#     mutate(lat = map_dbl(.$coord, ~as.numeric(str_split(., ",")[[1]][1])),
#            lon = map_dbl(.$coord, ~as.numeric(str_split(., ",")[[1]][2]))) %>% 
#     dplyr::select(prog_no, 
#                   program, 
#                   name, 
#                   census_tract, 
#                   students, 
#                   bldg_num, 
#                   grades, 
#                   type, 
#                   coord_type, 
#                   id, 
#                   lat,
#                   lon,
#                   am_bell, 
#                   pm_bell) %>% 
#     # five minutes before the bell
#     mutate(latest_arrival = as_hms((am_bell) - 5 * 60)) 
#   
#   # put latest arrival into the same format as am_bell (i.e. 8:00:00)
#   
#   # Initialize the latest valid departure time as the earliest possible
#   latest_departure_time <- NA
#   
#   school_join_object <- one_row_test %>% 
#     filter(coord_type != "home_tract_lat_lon") %>% 
#     dplyr::select(id, prog_no, bldg_num, name, type, grades, students, am_bell, pm_bell)
#   
#   # Loop to find the latest valid departure time
#   for (j in seq(from = max_search_time, to = 10, by = -10)) {
#     
#     departure_time <- as.POSIXct(paste(date, one_row_test[1,]$am_bell), format = "%Y-%m-%d %H:%M:%S") - j * 60 
#     
#     # Calculate travel time using r5r
#     travel_times <- travel_time_matrix(
#       r5r_core = r5r_core,
#       origins = one_row_test[one_row_test$coord_type == "home_tract_lat_lon", ],
#       destinations = one_row_test[one_row_test$coord_type == "school_lat_lon", ],
#       mode = mode,
#       departure_datetime = departure_time
#     ) %>% 
#       mutate(departure_time = departure_time,
#              on_time = travel_time_p50 <= as.numeric(difftime(as.POSIXct(paste(date, one_row_test[1,]$latest_arrival), format = "%Y-%m-%d %H:%M:%S"), departure_time, units = "mins")))
#     
#     # Check if the student can arrive on time
#     # if (nrow(travel_times) > 0 && travel_times$travel_time <= as.numeric(difftime(as.POSIXct(paste(date, one_row_test[1,]$am_bell), format = "%Y-%m-%d %H:%M:%S"), departure_time, units = "mins"))) {
#     if(travel_times$on_time == TRUE) {
#       latest_departure_time <- departure_time
#       
#       travel_times_list[[i]] <-
#         travel_times %>% 
#         left_join(school_join_object, by = c("to_id" = "id"))
#       
#       # print(latest_departure_time)
#     } else {
#       # If arrival time is too late, change the departure time to the next possible time in our sequence and start the loop again
#       print(departure_time)
#       # travel_times$departure_date <- latest_departure_time
#       break
#     }
#   }
# 
# # Store the result
# latest_departure_times[[i]] <- latest_departure_time
# 
# gc()
# 
# }
# 
# travel_times_from_student_student_home_tracts <- 
#   travel_times_list %>% 
#   bind_rows()  %>% 
#   mutate(arrival_time = departure_time + travel_time_p50 * 60) 
# 
# travel_times_from_student_student_home_tracts %>% 
#   summarise(mean_travel_time = weighted.mean(travel_time_p50, students, na.rm = T),
#             median_travel_time = median(travel_time_p50, students, na.rm = T),
#             sd_travel_time = sd(travel_time_p50, na.rm = T),
#             min_travel_time = min(travel_time_p50),
#             max_travel_time = max(travel_time_p50)) 
# 
# 


```

## Part II: Detailed Routing

Now that we basically know the parameters for our routing, we can model the actual routes students take to school. We'll use the 'detailed_itineraries' function from r5r to calculate the routes students take to school.

### Load data

```{r}
# 10/02
# latest_leaving_times <- read_rds(paste0("data/poa/", today, "/travel_times_df.rds")) %>%
#   mutate(departure_time = lubridate::with_tz(as.POSIXct(departure_time, format = "%Y-%m-%d %H:%M:%S"), "America/New_York"),
#          arrival_time = lubridate::with_tz(as.POSIXct(arrival_time, format = "%Y-%m-%d %H:%M:%S"), "America/New_York"))


latest_leaving_times_for_detailed_routing <- 
  travel_times_from_student_home_tracts %>% 
  mutate(prog_no = as.character(prog_no),
         from_id = str_remove_all(from_id, "x")
         # # convert arrival and departure times to EST
         # arrival_time = as.POSIXct(arrival_time, tz = "UTC") - hours(4),
         # departure_time = as.POSIXct(departure_time, tz = "UTC") - hours(4)
         ) %>% 
  left_join(student_home_tracts_23, 
            by = c(
              "from_id" = "census_tract",
              "prog_no" = "prog_no",
              "grades" = "grades",
              "name" = "name",
              "type" = "type",
              "students" = "students",
              "bldg_num" = "bldg_num"
            ))  %>% 
  rename(census_tract = from_id)

# latest_leaving_times_for_detailed_routing_car <-
#   latest_leaving_times_car %>% 
#   mutate(prog_no = as.character(prog_no),
#          from_id = str_remove_all(from_id, "x"),
#          # convert arrival and departure times to EST
#          arrival_time = as.POSIXct(arrival_time, tz = "UTC") - hours(4),
#          departure_time = as.POSIXct(departure_time, tz = "UTC") - hours(4)) %>% 
#   left_join(student_home_tracts_23, 
#             by = c(
#               "from_id" = "census_tract",
#               "prog_no" = "prog_no",
#               "grades" = "grades",
#               "name" = "name",
#               "type" = "type",
#               "students" = "students",
#               "bldg_num" = "bldg_num"
#             ))  %>% 
#   rename(census_tract = from_id)

```

### Transit

We'll start by modeling transit routes. In terms of actual code, this is actually a lot simpler than determining the latest departure time. We'll loop through each row in our data and use the 'detailed_itineraries' function to calculate the routes students take to school.

From there, we'll save each detailed route as a csv and load it back in to visualize the routes.

Output directory is hardcoded so just remember to change the output destination to the correct folder for the day you're modeling

```{r}
# Define modes of transit that we'll give our commuters access to 
mode <- c("WALK", "TRANSIT") 
i = 1


for(i in 1:nrow(latest_leaving_times_for_detailed_routing)){
  
  one_row <- latest_leaving_times_for_detailed_routing[i,]
  
  one_row_test <-
  one_row %>% 
    mutate(school_lat_lon = paste0(school_point_lat, ",", school_point_lon),
           home_tract_lat_lon = paste0(tract_point_lat, ",", tract_point_lon)) %>% 
    pivot_longer(
      cols = c("home_tract_lat_lon", "school_lat_lon"),
      names_to = "coord_type",
      values_to = "coord") %>% 
    mutate(id = ifelse(
      coord_type == "home_tract_lat_lon", 
      janitor::make_clean_names(census_tract), 
      str_replace_all(stringr::str_to_lower(program), " ", "_"))
      ) %>% 
    mutate(lat = map_dbl(.$coord, ~as.double(str_split(., ",")[[1]][1])),
           lon = map_dbl(.$coord, ~as.double(str_split(., ",")[[1]][2]))) %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
    mutate(geometry = st_make_valid(geometry)) %>% 
    # re-separate lat and lon
    mutate(lat = st_coordinates(geometry)[, 2],
           lon = st_coordinates(geometry)[, 1]) %>%
    dplyr::select(prog_no,
                  program,
                  name,
                  census_tract,
                  students,
                  bldg_num,
                  grades,
                  type,
                  coord_type,
                  id,
                  lat,
                  lon,
                  am_bell,
                  pm_bell)
  
  # make detailed routes directory inside of today's poa folder
  dir.create(here::here("data/poa", today, "detailed-routes"), showWarnings = F)

  df <- 
    detailed_itineraries(
      r5r_core,
      origins = one_row_test[1,],
      destinations = one_row_test[2,],
      mode = mode,
      departure_datetime = one_row$departure_time,
      shortest_path = T,
      output_dir = here::here("data/poa", today, "detailed-routes")
      )
  
  gc()
  
  Sys.sleep(0.1)
  
}

# test route file 
# detailed_route <- read_csv("/Users/greg.morton/Documents/baltimoreBanner/Projects/2024/high_school_transportation/data/poa/10-02-24/detailed-routes/from_x24510060100_to_stadium_school_m.csv")

library(readr)
from_x24510030100_to_stadium_school_m <- read_csv("data/poa/2025-01-17/detailed-routes/from_x24510030100_to_stadium_school_m.csv")
View(from_x24510030100_to_stadium_school_m)

```

# Read each csv file into a list of dataframes

```{r}
# read in the csv files

file_path <- paste0("data/poa/",today, "/detailed-routes")

file_paths <- list.files(here::here(file_path), full.names = T) %>% unique

df_list <- list()

df_list <- lapply(file_paths, read_csv)

```

# Combine the dataframes into one

```{r}
school_list <-
  df_list %>% 
  # mutate all from_id and to_id to character
  map(~mutate(.x, from_id = as.character(from_id), to_id = as.character(to_id))) %>%
  map(~mutate(.x, to_id = map_chr(to_id, ~str_remove_all(.x, "stop:")))) %>% 
  map(~mutate(.x, to_lat = as.character(to_lat), to_lon = as.character(to_lat))) %>% 
  map(~mutate(.x, departure_time = as.character(departure_time))) %>%
  map(~mutate(.x, total_distance = as.numeric(total_distance))) %>% 
  map(~mutate(.x, mode = as.character(mode))) %>% 
  map(~mutate(.x, wait = as.numeric(wait))) %>%
  bind_rows() %>% 
  group_by(from_id, to_id) %>%
  mutate(route = row_number()) 

geom <- school_list %>% 
  head(2) %>% 
  .$geometry

school_list %>% 
  st_as_sf(sf_column_name = "geometry")

school_list$geometry <- st_as_sfc(school_list$geometry)

school_list %>% 
  st_as_sf(crs = 4326) %>% 
  st_write(here::here("data/poa", today, "every_leg_every_trip.shp"))

school_list %>% 
  # filter(from_id == "x24510280302",
  #        to_id == "mer-vo_h") %>% 
  st_as_sf(crs = 4326) %>% 
  mapview(zcol = "mode") +
  mapview(baltimore_grid_sampled) +
  mapview(bmore_tracts_no_parks, alpha.regions = 0.2) +
  # add school shapes 
  mapview(hs_parcels_buffer, col.regions = "red") 

```

```{r}
# produce all trips file
all_trips <- 
  school_list %>%   
  mutate(route = as.character(route)) %>% 
  group_by(from_id, to_id) %>% 
  left_join(all_routes, by = c("route" = "id")) %>% 
  summarise(
    from_id = from_id[1],
    to_id = to_id[1],
    total_duration = total_duration[1],
    total_distance = total_distance[1],
    routes = list(route),
    short_names = list(short_name),
    modes = list(mode),
    trip_legs = max(segment),
    routes = list(route),
    bus_time = sum(segment_duration[mode == "BUS"], na.rm = T),
    bus_distance = sum(distance[mode == "BUS"], na.rm = T),
    walk_time = sum(segment_duration[mode == "WALK"], na.rm = T),
    walk_distance = sum(distance[mode == "WALK"], na.rm = T),
    subway_time = sum(segment_duration[mode == "SUBWAY"], na.rm = T),
    subway_distance = sum(distance[mode == "SUBWAY"], na.rm = T),
    light_rail_time = sum(segment_duration[mode == "TRAM"], na.rm = T),
    light_rail_distance = sum(distance[mode == "TRAM"], na.rm = T),
    wait = sum(wait, na.rm = T)
  ) %>% 
  ungroup() %>% 
  arrange(desc(trip_legs)) %>% 
  mutate(census_tract = str_remove_all(from_id, "x")) %>% 
  mutate(census_tract = as.numeric(census_tract)) %>%
  # select(from_id:to_id, everything()) %>% 
  # convert distance in meters to miles
  mutate(total_distance = total_distance * 0.000621371, 
         bus_distance = bus_distance * 0.000621371,
         walk_distance = walk_distance * 0.000621371,
         subway_distance = subway_distance * 0.000621371,
         light_rail_distance = light_rail_distance * 0.000621371) %>% 
  left_join(enrollment_23, by = c("to_id" = "program_join")) %>% 
  dplyr::select(census_tract, name, everything()) %>% 
  dplyr::select(-from_id) %>%
  mutate(census_tract = as.character(census_tract)) %>% 
  left_join(enrollment_by_tract, by = 
              c("census_tract" = "census_tract",
                "name" = "name"))  

  

# all_trips <- read_csv("/Users/greg.morton/Documents/baltimoreBanner/Projects/2024/high_school_transportation/data/poa/10-02-24/all_trips_10_2_24.csv")

# departure_times_10_02 <- 
#   read_csv("/Users/greg.morton/Documents/baltimoreBanner/Projects/2024/high_school_transportation/data/poa/10-02-24/departure_times_for_detailed_routing_10_2.csv") %>% 
#   mutate(departure_time = with_tz(departure_time, "America/New_York"),
#          arrival_time = with_tz(arrival_time, "America/New_York"), 
#          from_id = str_remove_all(from_id, "x")
#   ) %>% 
#   dplyr::select(from_id, to_id, departure_time, arrival_time, am_bell, pm_bell)
#  
# all_trips %>% 
#   mutate(routes = map_chr(routes, ~ paste(.x, collapse = ", ")),
#          short_names = map_chr(short_names, ~ paste(.x, collapse = ", ")),
#          short_names = map_chr(short_names, ~ paste(.x, collapse = ", ")),
#          modes = map_chr(modes, ~ paste(.x, collapse = ", "))) %>% 
#   left_join(departure_times_10_02, 
#             by = c("census_tract" = "from_id",
#                    "to_id" = "to_id")) %>% 
#   dplyr::select(census_tract, to_id, am_bell, departure_time, everything()) %>% 
#   write_csv("data/output/all_trips_update_10_02.csv")
```

```{r}


detailed_trip_summaries_every_school <- latest_leaving_times %>% 
  mutate(census_tract = str_remove_all(from_id, "x")) %>%
  mutate(prog_no = as.character(prog_no)) %>%
  left_join(all_trips, by = c("census_tract" = "census_tract",
                              "to_id" = "to_id",
                              "type" = "type",
                              "prog_no" = "prog_no",
                              "name" = "name",
                              "students" = "students",
                              "grades" = "grades"
                              )) %>% 
  dplyr::select(from_id, name, travel_time_p50, total_duration, everything()) 

# how many students on each route?
most_popular_routes <- 
  detailed_trip_summaries_every_school %>% 
  filter(!is.na(routes)) %>%
  mutate(routes = map(routes, ~unique(.x))) %>%
  clean_names() %>% 
  dplyr::select(routes, students, to_id, from_id) %>% 
  unnest() %>% 
  distinct(to_id, from_id, routes, .keep_all = T) %>% 
  group_by(routes) %>% 
    summarise(n = n(),
              total_students = sum(students)) %>% 
  left_join(all_routes, by = c("routes" = "id")) %>% 
  ungroup() %>% 
  rename(total_routes = n) %>% 
  # mutate(pct_of_trips = n/sum(n) * 100,
  #        sum = sum(pct_of_trips)
  # ) %>% 
  arrange(desc(total_students)) %>% 
  filter(!is.na(name)) %>% 
  write_csv("data/output/most_popular_routes_01-17.csv")

most_popular_routes %>% 
  summarise(total_routes = sum(n),
            total_students = sum(total_students))
  


# what are the most used routes to school? How many trips rely on each route?
most_used_routes <- 
  detailed_trip_summaries_every_school %>% 
  filter(!is.na(routes)) %>%
  mutate(routes = map(routes, ~unique(.x))) %>%
  clean_names() %>% 
  dplyr::select(routes) %>% 
  unnest() %>% 
  group_by(routes) %>% 
  summarise(n = n()) %>% 
  filter(!is.na(routes)) %>%
  left_join(all_routes, by = c("routes" = "id")) %>% 
  ungroup() %>% 
  arrange(desc(n)) %>%
  mutate(pct_of_trips = n/sum(n) * 100,
         sum = sum(pct_of_trips)
  )

```

```{r}
trip_summaries <- 
 detailed_trip_summaries_every_school %>% 
  # count instances of the string "WALK" in the modes column
  mutate(walk_instances = map_dbl(modes, ~sum(str_detect(.x, "WALK"))),
         transfers = ifelse(walk_instances -2 > 0, walk_instances - 2, 0),
         legs = map_dbl(modes, ~length(.x))
         ) %>% 
  summarise(
    enrollment_approx = sum(students, na.rm = T),
    # enrollment_actual = enrollment[1],
    bus_use_pct = sum(students[str_detect(modes, "BUS")])/enrollment_approx * 100,
    light_rail_use_pct = sum(students[str_detect(modes, "TRAM")])/enrollment_approx * 100,
    subway_use_pct = sum(students[str_detect(modes, "SUBWAY")])/enrollment_approx * 100,
    pct_students_greater_than_0_transfer = sum(students[transfers > 0])/enrollment_approx * 100,
    pct_students_greater_than_1_transfer = sum(students[transfers > 1])/enrollment_approx * 100,
    transfers = weighted.mean(transfers, students, na.rm = T),
    legs = weighted.mean(legs, students, na.rm = T),
    # average number of unique modes per trip
    avg_modes = weighted.mean(map_dbl(modes, ~length(unique(.x))), students, na.rm = T),
    # weighted averages of time and distance total, and by segment for all trips. Weighted by students
    total_duration = weighted.mean(total_duration, students, na.rm = T),
    total_distance = weighted.mean(total_distance, students, na.rm = T),
    median_walk_time = median(walk_time, students, na.rm = T),
    walk_time = weighted.mean(walk_time, students, na.rm = T),
    walk_distance = weighted.mean(walk_distance, students, na.rm = T),
    median_bus_time = median(bus_time, students, na.rm = T),
    mean_bus_time = weighted.mean(bus_time, students, na.rm = T),
    bus_distance = weighted.mean(bus_distance, students, na.rm = T),
    median_subway_time = median(subway_time, students, na.rm = T),
    mean_subway_time = weighted.mean(subway_time, students, na.rm = T),
    subway_distance = weighted.mean(subway_distance, students, na.rm = T),
    median_light_rail_time = median(light_rail_time, students, na.rm = T),
    mean_light_rail_time = weighted.mean(light_rail_time, students, na.rm = T),
    light_rail_distance = weighted.mean(light_rail_distance, students, na.rm = T)
  ) 
  # mutate(prog_no = as.numeric(prog_no)) 


# Longest trip to every school 
detailed_trip_summaries_every_school %>% 
  group_by(name, prog_no) %>% 
  # get max travel time for each group
  filter(travel_time_p50 == max(travel_time_p50)) 
# 
# write_csv(detailed_trip_summaries_every_school, "data/output/detailed_trip_summaries_every_school_09_11.csv")
# 
# write_csv(trip_summaries, "data/output/system_wide_trip_summaries_09_11.csv")
# 
# write_csv(most_used_routes, "data/output/most_used_routes_09_11.csv")


```

# What % of students travel longer than 45 minutes to school? Longer than one hour? Longer than 30 minutes on the bus?

```{r}

#citywide 
detailed_trip_summaries_every_school %>% 
  mutate(walk_instances = map_dbl(modes, ~sum(str_detect(.x, "WALK"))),
       transfers = ifelse(walk_instances -2 > 0, walk_instances - 2, 0),
       legs = map_dbl(modes, ~length(.x))
       ) %>% 
  summarise(
    enrollment_approx = sum(students, na.rm = T),
    avg_total_duration = weighted.mean(total_duration, students, na.rm = T),
    avg_total_total_distance = weighted.mean(total_distance, students, na.rm = T),
    pct_students_longer_than_30 = sum(total_duration > 30, na.rm = T)/n() * 100,
    pct_students_longer_than_45 = sum(total_duration > 45, na.rm = T)/n() * 100,
    pct_students_longer_than_60 = sum(total_duration > 60, na.rm = T)/n() * 100,
    pct_students_longer_than_30_bus = sum(bus_time > 30, na.rm = T)/n() * 100,
    pct_students_longer_than_45_bus = sum(bus_time > 45, na.rm = T)/n() * 100,
    pct_students_longer_than_30_walk = sum(walk_time > 30, na.rm = T)/n() * 100,
    pct_students_longer_than_45_walk = sum(walk_time > 45, na.rm = T)/n() * 100,
    pct_students_with_at_least_one_transfer = sum(transfers > 0, na.rm = T)/n() * 100,
    pct_students_with_more_than_one_transfer = sum(transfers > 1, na.rm = T)/n() * 100,
    pct_students_with_more_than_two_transfers = sum(transfers > 2, na.rm = T)/n() * 100
  ) 
  # write_csv("data/output/citywide_trip_time_summaries.csv")

# by school
detailed_trip_summaries_every_school %>% 
  group_by(name, prog_no) %>%
  mutate(walk_instances = map_dbl(modes, ~sum(str_detect(.x, "WALK"))),
       transfers = ifelse(walk_instances -2 > 0, walk_instances - 2, 0),
       legs = map_dbl(modes, ~length(.x))
       ) %>% 
  summarise(
    enrollment_approx = sum(students, na.rm = T),
    avg_total_duration = weighted.mean(total_duration, students, na.rm = T),
    avg_total_total_distance = weighted.mean(total_distance, students, na.rm = T),
    pct_students_longer_than_30 = sum(total_duration > 30, na.rm = T)/n() * 100,
    pct_students_longer_than_45 = sum(total_duration > 45, na.rm = T)/n() * 100,
    pct_students_longer_than_60 = sum(total_duration > 60, na.rm = T)/n() * 100,
    pct_students_longer_than_30_bus = sum(bus_time > 30, na.rm = T)/n() * 100,
    pct_students_longer_than_45_bus = sum(bus_time > 45, na.rm = T)/n() * 100,
    pct_students_longer_than_30_walk = sum(walk_time > 30, na.rm = T)/n() * 100,
    pct_students_longer_than_45_walk = sum(walk_time > 45, na.rm = T)/n() * 100,
    pct_students_with_at_least_one_transfer = sum(transfers > 0, na.rm = T)/n() * 100,
    pct_students_with_more_than_one_transfer = sum(transfers > 1, na.rm = T)/n() * 100,
    pct_students_with_more_than_two_transfers = sum(transfers > 2, na.rm = T)/n() * 100
  ) %>% 
  arrange(desc(pct_students_longer_than_30_bus)) 
  # write_csv("data/output/school_level_trip_time_summaries.csv")

# by tract 
# average_trip_by_tract <- 
  detailed_trip_summaries_every_school %>% 
  group_by(from_id) %>% 
  mutate(walk_instances = map_dbl(modes, ~sum(str_detect(.x, "WALK"))),
       transfers = ifelse(walk_instances -2 > 0, walk_instances - 2, 0),
       legs = map_dbl(modes, ~length(.x))
       ) %>% 
  summarise(
    enrollment_approx = sum(students, na.rm = T),
    avg_total_duration = weighted.mean(total_duration, students, na.rm = T),
    avg_total_total_distance = weighted.mean(total_distance, students, na.rm = T),
    pct_students_longer_than_30 = sum(total_duration > 30, na.rm = T)/n() * 100,
    pct_students_longer_than_45 = sum(total_duration > 45, na.rm = T)/n() * 100,
    pct_students_longer_than_60 = sum(total_duration > 60, na.rm = T)/n() * 100,
    pct_students_longer_than_30_bus = sum(bus_time > 30, na.rm = T)/n() * 100,
    pct_students_longer_than_45_bus = sum(bus_time > 45, na.rm = T)/n() * 100,
    pct_students_longer_than_30_walk = sum(walk_time > 30, na.rm = T)/n() * 100,
    pct_students_longer_than_45_walk = sum(walk_time > 45, na.rm = T)/n() * 100,
    pct_students_with_at_least_one_transfer = sum(transfers > 0, na.rm = T)/n() * 100,
    pct_students_with_more_than_one_transfer = sum(transfers > 1, na.rm = T)/n() * 100,
    pct_students_with_more_than_two_transfers = sum(transfers > 2, na.rm = T)/n() * 100
  ) %>% 
  arrange(desc(pct_students_longer_than_30_bus)) 
  # write_csv("data/output/tract_level_trip_time_summaries.csv")

# average_trip_by_tract %>% 
#   arrange(desc(pct_students_with_at_least_one_transfer))
# 
# bmore_tracts %>% 
#   left_join(average_trip_by_tract, by = c("GEOID" = "from_id")) %>% 
#   mapview(zcol = "pct_students_longer_than_45") 

```

# what time does the average student begin their commute? Median? Mode?

```{r}
departure_times_by_am_bell_time <- 
  detailed_trip_summaries_every_school %>% 
  group_by(am_bell) %>% 
  summarise(students = sum(students),
            mean_departure_time = mean(departure_time, na.rm = T),
            median_departure_time = median(departure_time, na.rm = T),
            earliest_departure_time = min(departure_time, na.rm = T),
            latest_departure_time = max(departure_time, na.rm = T)
            # most_common_departure_time = mode(departure_time, na.rm = T)
            ) 

# for all students 
departure_times_all_students <- 
  detailed_trip_summaries_every_school %>% 
  # group_by(am_bell) %>% 
  summarise(students = sum(students),
            mean_departure_time = mean(departure_time, na.rm = T),
            median_departure_time = median(departure_time, na.rm = T),
            earliest_departure_time = min(departure_time, na.rm = T),
            latest_departure_time = max(departure_time, na.rm = T)
            # most_common_departure_time = mode(departure_time, na.rm = T)
            )


tt_crosswalk <- travel_times_from_student_home_tracts %>% 
  group_by(to_id) %>% 
  summarise(name = name[1]) 
  # write_csv("data/output/school_names_to_id_crosswalk.csv")

hs_parcels_buffer %>% 
  left_join(tt_crosswalk, by = c("name" = "name")) %>% 
  st_write("data/output/hs_parcels_with_school_names.geojson")


```
